\section{Labour -- a derived demand}\label{sec:ch12sec1}

The value of labour springs from the value of its use, that is the value
placed upon goods and services that it produces -- product prices. The wage
is the price that equilibrates the supply and demand for a given type of
labour, and it reflects the value of that labour in production. Formally,
the demand for labour (and capital) is thus a \terminology{derived demand},
in contrast to being a `final' demand.

\begin{DefBox}
\textbf{Demand for labour}: a derived demand, reflecting the value of the output it produces.
\end{DefBox}

We must distinguish between the long run and the short run in our analysis
of factor markets. On the \textit{supply side} certain factors of production
are fixed in the short run. For example, the supply of radiologists can be
increased only over a period of years. While one hospital may be able to
attract radiologists from another hospital to meet a shortage, this does not
increase the supply in the economy as a whole.

On the \textit{demand side} there is the conventional difference between the
short and long run: In the short run some of a firm's factors of production,
such as capital, are fixed, and therefore the demand for labour differs from
when all factors are variable -- the long run.

\newhtmlpage

\subsection*{Demand in the short run}

Table~\ref{table:srprodlabourdemand} contains information from the example
developed in Chapter~\ref{chap:prodcost}. It can be used to illustrate how a
firm reacts in the short run to a change in an input price, or to a change
in the output price. The response of a producer to a change in the wage rate
constitutes a demand function for labour -- a schedule relating the quantity
of the input demanded to different input prices. The output produced by the
various numbers of workers yields a marginal product curve, whose values are
stated in column 3. The marginal product of labour, $MP_{L}$, as developed
in Chapter~\ref{chap:prodcost}, is the additional output resulting from one
more worker being employed, while holding constant the other (fixed)
factors. But what is the \textit{dollar value} to the firm of an additional
worker? It is the additional \textit{value of output} resulting from the
additional employee -- the price of the output times the worker's marginal
contribution to output, his $MP$. We term this the \terminology{value of the
marginal product}.

\begin{DefBox}
The \textbf{value of the marginal product} is the marginal product multiplied by the price of the good produced.
\end{DefBox}

\begin{Table}{caption={Short-run production and labour demand \label{table:srprodlabourdemand}},description={Each unit of labour costs \$1,000; output sells at a fixed price of \$70 per unit.},descwidth={26em}}
\begin{tabu} to \linewidth {|X[1,c]X[1,c]X[1,c]X[1.5,c]X[1.5,c]|} \hline 
\rowcolor{rowcolour}	\textbf{Workers} & \textbf{Output} & \textbf{$MP_L$} & \textbf{$VMP_L=MP_L\times P$} & \textbf{Marginal profit = ($VMP_L-$wage) } \\[-0.05em]
\rowcolor{rowcolour}\textbf{(1)}	&	\textbf{(2)}	&	\textbf{(3)}	&	\textbf{(4)}	&	\textbf{(5)}	\\	\hline
						0 & 0	&		&		&			\\
\rowcolor{rowcolour}	1 & 15	& 15	& 1050	& 150		\\
						2 & 40	& 25	& 1750	& 750		\\
\rowcolor{rowcolour}	3 & 70	& 30	& 2100	& 1100		\\ 
						4 & 110 & 40	& 2800	& 1800		\\
\rowcolor{rowcolour}	5 & 145 & 35	& 2450	& 1450		\\
						6 & 175 & 30	& 2100	& 1100		\\
\rowcolor{rowcolour}	7 & 200 & 25	& 1750	& 750		\\
						8 & 220 & 20	& 1400	& 400		\\
\rowcolor{rowcolour}	9 & 235 & 15	& 1050	& 50		\\
						10& 240 & 5		& 350	& negative	\\ \hline 
\end{tabu}
\end{Table}

\newhtmlpage



In this example the $MP_{L}$ first rises as more labour is employed, and
then falls. With each unit of output selling for \$70 the value of the
marginal product of labour ($VMP_{L}$) is given in column 4. The first
worker produces 15 units each week, and since each unit sells for a price of
\$70, his production value to the firm is \$1,050 $(=\$70\times 15)$. A
second worker produces 25 units, so his value to the firm is \$1,750, and so
forth. If the weekly wage of each worker is \$1,000 then the firm can
estimate its marginal profit from hiring each additional worker. This is the
difference between the value of the marginal product and the wage paid, and
is given in the final column of the table.

It is profitable to hire more workers as long as the cost of an extra worker
is less than the $VMP_{L}$. The equilibrium amount of labour to employ is
therefore 9 units in this example. If the firm were to hire one more worker
the contribution of that worker to its profit would be negative 
$(\$350-\$1,000)$, and if it hired one worker less it would forego the
opportunity to make an additional profit of \$50 on the 9th unit 
$(\$1,050-\$1,000)$.


\textit{Profit maximizing hiring rule:}
\begin{itemize}
\item \textit{If the $VMP_L$ of next worker $>$ wage, hire more labour.}

\item \textit{If the $VMP_L$ $<$ wage, hire less labour.}
\end{itemize}

To this point we have determined the profit maximizing amount of labour to
employ when the output price and the wage are given. However, a demand
function for labour reflects the demand for labour at many different wage
rates, just as the demand function for any product reflects the quantity demanded at
various prices. Accordingly, suppose the wage rate is \$1,500 per week
rather than \$1,000. The optimal amount of labour to employ in this case is
determined in exactly the same manner: Employ the amount of labour where its
contribution is marginally profitable. Clearly the optimal amount to employ
is 7 units: The value of the seventh worker to the firm is \$1,750 and the
value of the eighth worker is \$1,400. Hence it would not be profitable to
employ the eighth, because his marginal contribution to profit would be
negative. Following the same procedure we could determine the optimal amount
of labour to employ at any wage. This implies that the $VMP_{L}$ function is
the demand for labour function because it determines the most profitable
amount of labour to employ at any wage.

\newhtmlpage

The optimal amount of labour to hire is illustrated in Figure~\ref{fig:demandforlabour}.
The wage and $VMP_L$ curves come from Table~\ref{table:srprodlabourdemand}.
The $VMP_L$ curve has an upward sloping segment,
reflecting increasing productivity, and then a regular downward slope as
developed in Chapter~\ref{chap:prodcost}. At employment levels where the 
$VMP_L$ is greater than the wage additional labour should be employed. But
when the $VMP_L$ falls below the wage rate employment should stop. If labour
is divisible into very small units, the optimal employment decision is where
the $MP_L$ function intersects the wage line.

% Figure 12.1
\input{figures/ch12/ch12fig1}

\newhtmlpage

Figure~\ref{fig:demandforlabour} also illustrates what happens to hiring
when the output price changes. Consider a reduction in its price to \$50
from \$70. The profit impact of such a change is negative because the value
of each worker's output has declined. Accordingly, the demand curve must
reflect this by shifting inward (down), as in the figure. At various wage
rates, less labour is now demanded. The new $VMP_{L}$ schedule can be
derived in Table~\ref{table:srprodlabourdemand} as before: It is the $MP_{L}$
schedule multiplied by the lower value (\$50) of the final good.

In this example the firm must be seen as a perfect competitor in the output
market, because the price of the good it produces is fixed. It can produce
and sell more of the good without this having an impact on the price of the
good in the marketplace. Where the firm is not a perfect competitor it faces
a declining $MR$ function. In this case the value of the $MP_{L}$ is the
product of $MR$ and $MP_{L}$ rather than $P$ and $MP_{L}$. To distinguish
the different output markets we use the term \terminology{marginal
\textit{revenue} product of labour} ($MRP_{L}$) when the demand for the
output slopes downward. But the optimizing principle remains the same: The
firm should calculate the value of each additional unit of labour, and hire
up to the point where the additional revenue produced by the worker exceeds
or equals the additional cost of that worker.

\begin{DefBox}
The \textbf{marginal revenue product of labour} is the additional revenue generated by hiring one more unit of labour where the marginal revenue declines.
\end{DefBox}

\newhtmlpage

\subsection*{Demand in the long run}

In Chapter~\ref{chap:prodcost} we proposed that firms choose their factors
of production in accordance with cost-minimizing principles. In producing a
specific output, firms choose the least-cost combination of labour and plant
size. But how is this choice affected when the price of labour or capital
changes? While adjustment to price changes may require a long period of
time, we know that if one factor becomes more (less) expensive, the firm
will likely change the mix of capital and labour away from (towards) that
factor. For example, when the accuracy and prices of production robots began
to fall in the nineteen nineties, auto assemblers reduced their labour and
used robots instead. When computers and computer software improved and
declined in price, clerical workers were replaced by computers that were
operated by accountants. But such adjustments and responses do not occur
overnight.

In the short run a higher wage increases costs, but the firm is constrained
in its choice of inputs by a fixed plant size. In the long run, a wage
increase will induce the firm to use relatively more capital than when
labour was less expensive in producing a given output. But despite the new
choice of inputs, a rise in the cost of any input must increase the total
cost of producing any output.

A change in the price of any factor has two impacts on firms: In the first
place they will \textit{substitute} away from the factor whose price
increases; second, there will be an impact on output and a change in the
price of the final good it produces. Since the cost structure increases when
the price of an input rises, the supply curve in the market for the good
must reflect this -- any given output will now be supplied at a higher
price. With a downward sloping demand, this shift in supply must increase
the price of the good and reduce the amount sold. This second effect can be
called an \textit{output effect}.

\newhtmlpage

\subsection*{Monopsony}

Some firms may have to pay a higher wage in order to employ more workers.
Think of Hydro Quebec building a dam in Northern Quebec. Not every hydraulic
engineer would be equally happy working there as in Montreal. Some engineers
may demand only a small wage premium to work in the North, but others will
demand a high premium. If so, Hydro Quebec must pay a higher wage to attract
more workers -- it faces an upward sloping supply of labour curve. Hydro
Quebec is the sole buyer in this particular market and is called a %
\terminology{monopsonist} -- a single buyer. Our general optimizing
principle governing the employment of labour still holds, even if we have
different names for the various functions: Hire any factor of production up
to the point where the cost of an additional unit equals the value generated
for the firm by that extra worker. The essential difference here is that
when a firm faces an upward sloping labour supply it will have to pay more
to attract additional workers and \textit{also pay more to its existing
workers}. This will impact the firm's willingness to hire additional workers.

\begin{DefBox}
A \textbf{monopsonist} is the sole buyer of a good or service and faces an upward-sloping supply curve.
\end{DefBox}

\subsection*{Firm versus industry demand}

The demand for labour within an industry, or sector of the economy, is
obtained from the sum of the demands by each individual firm. It is
analogous to the goods market, but with a subtle difference. In Chapter~\ref{chap:classical} 
we obtained a market demand by summing individual demands horizontally. The same
could be done here: At lower (or higher) wages, each firm will demand
more (or less) labour. However, if all firms employ more labour in order to
increase their output, the price of the output will likely decline. This in
turn will moderate the demand for labour -- it is slightly less valuable now
that the price of the output it produces has fallen. This is a subtle point,
and we can reasonably think of the demand for labour in a given sector of
the economy as the sum of the demands on the part of the employers in that
sector.
