\chapter{Government}\label{chap:government}

\begin{topics}
	\textbf{In this chapter we will explore:}
	\begin{description}
		\item[\ref{sec:ch14sec1}] Market failure and the role of government
		\item[\ref{sec:ch14sec2}] Fiscal federalism in Canada
		\item[\ref{sec:ch14sec3}] Federal-provincial relations and powers
		\item[\ref{sec:ch14sec4}] Redistribution to individuals
		\item[\ref{sec:ch14sec5}] Regulatory activity and competition policy
	\end{description}
\end{topics}

Governments have a profound impact on economies. The economies of
Scandinavia are very different from those in North America. North and South
Korea are night and day, even though they were identical several decades
ago. Canada and Argentina were very similar in the early decades of the
twentieth century. Both had abundant space, natural resources and migrants.
Today Canada is one of the most prosperous economies in the world, while
Argentina struggles as a middle-income economy.

Governments are not peripheral to the marketplace. They can facilitate or
hinder the operation of markets. They can ignore poverty or implement
policies to support those on low income and those who are incapacitated.
Governments can treat the economy as their fiefdoms, as has been the case
for decades in many underdeveloped economies. By assuming the role of
warlords, local governors inhibit economic development, because the fruits
of investment and labour are subject to capture by the ruling power elite.

In Canada we take for granted the existence of a generally benign government
that serves the economy, rather than one which expects the economy to serve
it. The separation of powers, the existence of a constitution, property
rights, a police force, and a free press are all crucial ingredients in the
mix that ferments economic development and a healthy society.

The analysis of government is worthy not just of a full course, but a full
program of study. Accordingly, our objective in this chapter must be
limited. We begin by describing the various ways in which markets may be
inadequate and what government can do to remedy these deficiencies. Next we
describe the size and scope of government in Canada, and define the sources
of government revenues. On the expenditure side, we emphasize the
redistributive and transfer roles that are played by Canadian governments.
Tax revenues, particularly at the federal level, go predominantly to
transfers rather than the provision of goods and services. Finally we
examine how governments seek to control, limit and generally influence the
marketplace: How do governments foster the operation of markets in Canada?
How do they attempt to limit monopolies and cartels? How do they attempt to
encourage the entry of new producers and generally promote a market
structure that is conducive to competition, economic growth and consumer
well-being?