\section{Fiscal federalism: Taxing and spending}\label{sec:ch14sec2}

Canada is a federal state, in which the federal, provincial and municipal
governments exercise different powers and responsibilities. In contrast,
most European states are unitary and power is not devolved to their regions
to the same degree as in Canada or the US or Australia. Federalism confers
several advantages over a unitary form of government where an economy is
geographically extensive, or where identifiable differences distinguish one
region from another: Regions can adopt different policies in response to the
expression of different preferences by their respective voters; smaller
governments may be better at experimentation and the introduction of new
policies than large governments; political representatives are `closer' to
their constituents.

Despite these advantages, the existence of an additional level of government
creates a tension between these levels. Such tension is evident in every
federation, and federal and provincial governments argue over the
appropriate division of taxation powers and revenue-raising power in
general. For example, how should the royalties and taxes from oil and gas
deposits offshore be distributed -- to the federal government or a
provincial government?

In Canada, the federal government collects more in tax revenue than it
expends on its own programs. This is a feature of most federations. The
provinces simultaneously face a shortfall in their own revenues relative to
their program expenditure requirements. The federal government therefore
redistributes, or transfers, funds to the provinces so that the latter can
perform their constitutionally-assigned roles in the economy. The fact that
the federal government bridges this fiscal gap gives it a degree of power
over the provinces. This influence is commonly termed \terminology{federal
	spending power}.

\begin{DefBox}
	\textbf{Spending power} of a federal government arises when the federal government can influence lower level governments due to its financial rather than constitutional power.
\end{DefBox}

\newhtmlpage

The principal revenue sources for financing government activity are given in
the pie chart in Figure~\ref{fig:fedrev} for the fiscal year 2010/11, and
the expenditure of these revenues is broken down in Figure~\ref{fig:fedexp}.
Further details are accessible at the Department of Finance's web site,
though students should be aware that government accounts come in many forms.
Total revenues for this fiscal year amounted to \$237b.

% Figure 14.2
\input{figures/ch14/ch14fig2}

%Figure 14.3
\input{figures/ch14/ch14fig3}

\newhtmlpage

The federal and provincial governments each transfer these revenues to
individuals and other levels of government, supply goods and services
directly, and also pay interest on accumulated borrowings -- the national
debt or provincial debt.

Provincial and local governments supply more goods and services than the
federal government -- health care, drug insurance, education and welfare are the
responsibility of provincial and municipal governments. In contrast,
national defence, the provision of main traffic arteries, Corrections Canada
and a variety of transfer programs to individuals -- such as Employment
Insurance, Old Age Security and the Canada Pension Plan -- are federally
funded. The greater part of federal revenues goes towards \textit{transfers}
to individuals and provincial governments, as opposed to the supply of goods
and services.