\section{Federal-provincial fiscal relations}\label{sec:ch14sec3}

The federal government transfers revenue to the provinces using three main
programs: Equalization, the Canada Social Transfer and the Canada Health
Transfer. Each of these has a different objective. Equalization aims to
reduce fiscal disparities among the provinces; The Canada Social Transfer
(CST) is for educational and Social Assistance (`welfare') expenditures; The
Canada Health Transfer (CHT) performs the same function for health.

\subsection*{Equalization}

\begin{quote}
	\textit{``Parliament and the Government of Canada are committed to the
		principle of making equalization payments to ensure that provincial
		governments have sufficient revenues to provide reasonably comparable levels
		of public service at reasonably comparable levels of taxation.''}
\end{quote}

This statement, from Section 36(2) of the Constitution Act of 1982, defines
the purpose of Equalization. Equalization payments are unconditional --
receiving provinces are free to spend the funds on public services according
to their own priorities, or even use the revenue to reduce their provincial
taxes. Payments are calculated according to a formula that ensures those
provinces with revenue-raising ability, or fiscal capacity, below a
threshold or `standard' receive payments from the federal government to
bring their capacity up to that standard.

Equalization has gone through very many changes in the several decades of
its existence. Its current status reflects the 2006 recommendations of an
Expert Panel set up by the federal government. The fiscal capacity of a
province is measured by its ability to raise revenues from five major
sources: Personal and business income taxes, sales taxes, property taxes,
and natural resources. This ability is then compared to the ability of all
of the provinces combined to raise revenue; if a difference or shortfall
exists, the federal government transfers revenue accordingly, with the
amount determined by both the population of the province and the magnitude
of its per-person shortfall.

The program currently transfers about \$16b per annum. The recipiency status
of some provinces varies from year to year. Variation in energy prices and
energy-based government revenues are the principal cause of this. British
Columbia, Alberta, Saskatchewan and Ontario tend to receive little or zero.
Manitoba, Quebec and the Atlantic Provinces have been the major recipient
provinces. Quebec receives the largest single amount -- more than half the
total allocation, on account of its population size. Details are available
at the federal government's Department of Finance web site.

\newhtmlpage

\subsection*{The Canada Social Transfer and the Canada Health Transfer}

The CST is a block transfer to provinces in support of post-secondary
education, Social Assistance and social services more generally. The CST
came into effect in 2004. Prior to that date it was integrated with the
health component of federal transfers in a program titled the Canada Health
and Social Transfer (CHST). The objective of the separation was to increase
the transparency and accountability of federal support for health while
continuing to provide funding for other objectives. The CHT is the other
part of the unbundled CHST: It provides funding to the provinces for their
health expenditures.

The CST and CHT funding comes in two parts: A cash transfer and tax
transfer. A tax transfer essentially provides the same support as a cash
transfer of equal value; it just comes in a different form. In 1977 the
federal government agreed with provincial and territorial governments to
reduce federal personal and corporate tax rates in order to permit the
provincial governments to increase the corresponding provincial rates. The
net effect was that the federal government got less tax revenue and the
provinces got more. And to this day, the federal and provincial governments
keep a record of the implied tax transfers that arise from this long-ago
agreement. This is the tax transfer component of the CST and the CHT.

The CST support is allocated to provinces and territories on an equal
per-capita basis to ensure equal support for all Canadians regardless of
their place of residence. The CHT is distributed likewise, and it requires
the provinces to abide by the federally-legislated \textit{Canada Health Act}, 
which demands that provincial health coverage be comprehensive, universal,
portable, accessible and publicly administered.

Approximately \$35b is currently transferred under the CHT and \$13b under
the CST, in cash. Health care and health expenditures are a core issue of
policy at all levels of government on account of the envisaged growth in
these expenditures that will inevitably accompany the aging of the
baby-boomers.