\section{Government-to-individual transfers}\label{sec:ch14sec4}

Many Canadians take pride in Canada's extensive `social safety net' that
aims to protect individuals from misfortune and the reduction of income in
old age. Others believe it is too generous. While it is more supportive than
the safety net in the US, the Canadian safety net is no more protective than
the nets of the developed economies of the European Union. The extent of
such support depends in large measure upon the degree to which governments
are willing to impose, and individuals are willing to pay, higher or lower
tax rates. The major elements of this umbrella of programs are the following.

The \textit{Canada and Quebec Pension Plans (C/QPP)} are funded from
the contributions of workers and their employers. Contributions form 9.9\%
of an individual's earnings up to a maximum of approximately \$52,000. The
Canada and Quebec components of the plan operate very similarly, but are
managed separately. Contributions to the plans from workers and their
employers are, for the most part, paid immediately to retired workers. Part
of the contributions is invested in a fund.

The objective of the plans is to ensure that some income is saved for
retirement. Many individuals are not very good at planning -- they
constantly postpone the decision to save, so the state steps in and requires
them to save. An individual contributing throughout a full-time working
lifecycle can expect an annual pension of about \$12,000. The objective of
the plan is to provide a minimum level of retirement income, not an income
that will see individuals live in great comfort. The C/QPP plans have
contributed greatly to the reduction of poverty among the elderly since
their introduction in the mid-sixties.

The aging of the baby-boom generation -- that very large cohort born in the
late forties through to the early sixties -- means that the percentage of
the population in the post-65 age group will increase radically, beginning
about 2015. To meet this changing demographic, the federal and provincial
governments reshaped the plans in the late nineties -- primarily by
increasing contributions, in order to build up a fund that will support the
aged in the following decades. 

Some provincial governments are presently concerned that many
middle-earning Canadians are saving insufficiently for their retirement, in
the sense that their standard of living will likely drop at the point of
retirement. The federal government of Stephen Harper decided not to expand
the CPP. Expanding the CPP would force more households to contribute more to
that Plan. In contrast, the Ontario government of Premier Wynne has gone
ahead with its own pension legislation. This will require workers to
contribute to the Ontario plan on earnings up to \$90,000. Full benefits
will be paid to contributors after a four-decade contribution period.  

\textit{Old Age Security (OAS)}, the \textit{Guaranteed Income
	Supplement (GIS)} and the \textit{Spousal Allowance (SPA)} together
form the second support leg for the retired. OAS is a payment made
automatically to individuals once they attain the age of 65. The GIS is an
additional payment made only to those on very low incomes -- for example,
individuals who have little income from their C/QPP or private pension
plans. The SPA, which is payable to the spouse or survivor of an OAS
recipient, accounts for a small part of the sums disbursed.

\newhtmlpage

The payments for these plans come from the general tax revenues of the
federal government. Unlike the C/QPP, the benefits received are not related
to the contributions that an individual makes over the working lifecycle.
This program has also had a substantial impact on poverty reduction among
the elderly.

\textit{Employment Insurance (EI)} and \textit{Social Assistance (SA)} 
are designed to support, respectively, the unemployed and those with no
other source of income. Welfare is the common term used to describe SA.
Expenditures on EI and SA are strongly cyclical. At the trough of an
economic cycle the real value of expenditures on these programmes greatly
exceeds expenditures at the peak of the cycle. Unemployment in Canada rose
above 8\% in 2009, and payments to the unemployed and those on welfare
reflected this dire state. The strongly cyclical pattern of the cost of
these programs reflects the importance of a healthy job market:
Macroeconomic conditions have a major impact on social programme
expenditures.

EI is funded by contributions from employees and their employers. As of 2015
employees pay 1.88\% of their earnings and their employer pays 1.4 times
this amount. EI is called an insurance programme, but in reality it is much
more than that. Certain groups systematically use the program more than
others -- those in seasonal jobs, those in rural areas and those in the
Atlantic Provinces, for example. Accordingly, using the terminology of
Chapter~\ref{chap:firminvestorcapital}, it is not everywhere an actuarially
`fair' insurance programme. Benefits payable to unemployed individuals may
also depend on their family size, in addition to their work history. While
most payments go in the form of `regular' benefits to unemployed
individuals, the EI program also sponsors employee retraining, family
benefits that cover maternity and paternity leave, and some other specific
target programs for the unemployed.

Social Assistance is provided to individuals who are in serious need of
financial support -- having no income and few assets. Provincial governments
administer SA, although the cost of the programme is partly covered by
federal transfers through the Canada Social Transfer. The nineteen nineties
witnessed a substantial tightening of regulations virtually across the whole
of Canada. Access to SA benefits is now more difficult, and benefits have
fallen in real terms since the late nineteen eighties.

Welfare dependence peaked in Canada in 1994, when 3.1 million individuals
were dependent upon support. As of 2015, the total is a little more than one
half of this, on account of more stringent access conditions, reduced
benefit levels and an improved job market. Some groups in Canada believe
that benefits should be higher, others believe that making welfare too
generous provides young individuals with the wrong incentives in life, and
may lead them to neglect schooling and skill development.

\textit{Workers Compensation} supports workers injured on the job.
Worker/employer contributions and general tax revenue form the sources of
program revenue, and the mix varies from province-to-province. In contrast
to the macro-economy-induced swings in expenditures that characterize SA and
EI since the early nineties, expenditures on Worker's Compensation have
remained relatively constant.

\newhtmlpage

The major remaining pillar in Canada's social safety net is the group of 
payments and tax credits aimed at supporting children: The \textit{Canada 
Child Tax Benefit (CCTB)}, the \textit{Universal Child Care Benefit} and the
\textit{National Child Benefit Supplement}. These programs together cost the
government in the neighbourhood of \$16b. The program group has
evolved and been enriched over the last two decades, partly with the
objective of reducing poverty among households with children, and partly
with a view to helping parents receiving SA to transit back to the labour
market.

\begin{ApplicationBox}{caption={Government debts, deficits and transfer \label{app:govdebtdeftran}}}
	Canada's expenditure and tax policies in the nineteen seventies and eighties led to the accumulation of large government debts, as a result of running fiscal deficits most of those years. By the mid-nineties the combined federal and provincial debt reached 100\% of GDP, with the federal debt accounting for two thirds of this amount. This ratio was perilously high: Interest payments absorbed a large fraction of annual revenues, which in turn limited the ability of the government to embark on new programs or enrich existing ones. Canada's debt rating on international financial markets declined.

	In 1995, Finance Minister Paul Martin addressed this problem, and over the following years program spending was pared back. Ultimately, the economy expanded and by the end of the decade the annual deficits at the federal level were eliminated.

	As of 2007 the ratio of combined federal and provincial debts stood at just over 60\% of GDP, however, the Great Recession of 2008 and following years saw all levels of government experience deficits, with the result that this ratio now is in excess of 80\%. The federal budget resumed a state of approximate balance by 2015, while provinces in total incurred deficits of approximately 20 billion dollars. 
\end{ApplicationBox}