\section{Regulation and competition policy}\label{sec:ch14sec5}

\subsection*{Goals and objectives}

The goals of competition policy are relatively uniform across developed
economies: The promotion of domestic competition; the development of new
ideas, new products and new enterprises; the promotion of efficiency in the
resource-allocation sense; the development of manufacturing and service
industries that can compete internationally.

In addition to these economic objectives, governments and citizens frown
upon monopolies or monopoly practices if they lead to an undue \textit{%
	concentration of political power}. Such power can lead to a concentration of
wealth and influence in the hands of an elite.

Canada's regulatory body is the \textit{Competition Bureau}, whose activity
is governed primarily by the \textit{Competition Act} of 1986. This act
replaced the \textit{Combines Investigation Act}. The \textit{Competition
	Tribunal} acts as an adjudication body, and is composed of judges and non-
judicial members. This tribunal can issue orders on the maintenance of
competition in the marketplace. Canada has had anti-combines legislation
since 1889, and the act of 1986 is the most recent form of such legislation
and policy. The Competition Act does not forbid monopolies, but it does rule
as unlawful the \textit{abuse} of monopoly power. Canada's competition
legislation is aimed at anti-competitive practices, and a full description
of its activities is to be found on its website at \href{www.competitionbureau.gc.ca}{www.competitionbureau.gc.ca}.
Let us examine some of these proscribed
policies.

\newhtmlpage

\subsection*{Anti-competitive practices}

Anti-competitive practices may either limit entry into a sector of the
economy or force existing competitors out. In either case they lead to a
reduction in competition.

\textit{Mergers} may turn competitive firms into a single organization
with excessive market power. The customary justification for mergers is that
they permit the merged firms to achieve scale economies that would otherwise
be impossible. Such scale economies may in turn result in lower prices in
the domestic or international market to the benefit of the consumer, but may
alternatively reduce competition and result in higher prices. Equally
important in this era of global competition is the impact of a merger on a
firm's ability to compete internationally.

Mergers can be of the horizontal type (e.g. two manufacturers of pre-mixed
concrete merge) or vertical type (a concrete manufacturer merges with a
cement manufacturer). While mergers have indeed the potential to reduce
domestic competition, this is less true today than in the past, when
international trade was more restricted.

\terminology{Cartels} aim to restrict output and thereby increase profits.
These formations are almost universally illegal in individual national
economies. At the same time governments are rarely worried if their own
national firms participate in international cartels.

While cartels are one means of increasing prices, \terminology{price
	discrimination} is another, as we saw when studying monopoly behaviour. For
example, if a concrete manufacturer makes her product available to large
builders at a lower price than to small-scale builders -- perhaps because
the large builder has more bargaining power -- then the small builder is at a
competitive disadvantage in the construction business. If the small firm is
forced out of the construction business as a consequence, then competition
in this sector is reduced.

We introduced the concept of predatory pricing in Chapter~\ref{chap:imperfectcompetition}. 
\terminology{Predatory pricing} is a practice that is aimed at driving out
competition by artificially reducing the price of one product sold by a
supplier. For example, a dominant nationwide transporter could reduce price
on a particular route where competition comes from a strictly local
competitor. By `subsidizing' this route from profits on other routes, the
dominant firm could undercut the local firm and drive it out of the market.

\begin{DefBox}
	\textbf{Predatory pricing} is a practice that is aimed at driving out competition by artificially reducing the price of one product sold by a supplier.
\end{DefBox}

\newhtmlpage

Suppliers may also \terminology{refuse to deal}. If the local supplier of
pre-mixed concrete refuses to sell the product to a local construction firm,
then the ability of such a downstream firm to operate and compete may be
compromised. This practice is similar to that of \terminology{exclusive
	sales} and \terminology{tied sales}. An exclusive sale might involve a large
vegetable wholesaler forcing her retail clients to buy \textit{only} from
this supplier. Such a practice might hurt the local grower of aubergines or
zucchini, and also may prevent the retailer from obtaining \textit{some} of
her vegetables at a lower price or at a higher quality elsewhere. A \terminology{tied
	sale} is one where the purchaser must agree to purchase a \textit{bundle} of
goods from a supplier.

\begin{DefBox}
	\textbf{Refusal to deal}: an illegal practice where a supplier refuses to sell to a purchaser.
	
	\textbf{Exclusive sale}: where a retailer is obliged (perhaps illegally) to purchase all wholesale products from a single supplier only.
	
	\textbf{Tied sale}: one where the purchaser must agree to purchase a bundle of goods from a supplier.
\end{DefBox}

\terminology{Resale price maintenance} involves the producer requiring a
retailer to sell a product at a specified price. This practice can hurt
consumers since they cannot `shop around'. In Canada, we frequently
encounter a `manufacturer's suggested retail price' for autos and durable
goods. But since these prices are not \textit{required}, the practice
conforms to the law.

\begin{DefBox}
	\textbf{Resale price maintenance} is an illegal practice wherein a producer requires sellers to maintain a specified price.
\end{DefBox}

\newhtmlpage

\terminology{Bid rigging} is an illegal practice in which normally
competitive bidders conspire to fix the awarding of contracts or sales. For
example, two builders, who consider bidding on construction projects, may
decide that one will bid seriously for project X and the other will bid
seriously on project Y. In this way they conspire to reduce competition in
order to make more profit.

\begin{DefBox}
	\textbf{Bid rigging} is an illegal practice in which bidders (buyers) conspire to set prices in their own interest.
\end{DefBox}

\textit{Deception and dishonesty in promoting products} can either
short-change the consumer or give one supplier an unfair advantage over
other suppliers.

\subsection*{Enforcement}

The Competition Act is enforced through the Competition Bureau in a variety
of ways. Decisions on acceptable business practices are frequently reached
through study and letters of agreement between the Bureau and businesses. In
some cases, where laws appear to have been violated, criminal proceedings
may follow.

\newhtmlpage

\subsection*{Regulation, deregulation and privatization}

The last three decades have witnessed a significant degree of privatization
and deregulation in Canada, most notably in the transportation,
communication and energy sectors. Modern deregulation in the US began with
the passage of the \textit{Airline Deregulation Act} of 1978, and was
pursued with great energy under the Reagan administration in the eighties.
The Economic Council of Canada produced an influential report in 1981,
titled ``Reforming Regulation,'' on the
impact of regulation and possible deregulation of specific sectors. The
Economic Council proposed that regulation is some sectors was inhibiting
competition, entry and innovation. As a consequence, the interests of the
consumer were in danger of becoming secondary to the interests of the
suppliers.

Telecommunications provision, in the era when the telephone was the main
form of such communication, was traditionally viewed as a natural monopoly.
The Canadian Radio and Telecommunications Commission (CRTC) regulated its
rates. The industry has developed dramatically in the last decade with the
introduction of satellite-facilitated communication, the internet,
multi-purpose cable networks, cell phones and service integration. With such
rapid development in new technologies and modes of delivery the CRTC has
been playing catch-up to these new realities.

Transportation, in virtually all forms, has been deregulated in Canada since
the nineteen eighties. Railways were originally required to subsidize the
transportation of grain under the \textit{Crow's Nest Pass} rate structure.
But the subsidization of particular markets requires an excessive rate
elsewhere, and if the latter markets become subject to competition then a
competitive system cannot function. This structure, along with many other
anomalies, was changed with the passage of the \textit{Canada Transportation
	Act} in 1996.

Trucking, historically, has been regulated by individual provinces. Entry
was heavily controlled prior to the federal \textit{National Transportation
	Act} of 1987, and subsequent legislation introduced by a number of
provinces, have made for easier entry and a more competitive rate structure.

Deregulation of the airline industry in the US in the late seventies had a
considerable influence on thinking and practice in Canada. The Economic
Council report of 1981 recommended in favour of easier entry and greater
fare competition. These policies were reflected in the 1987 National
Transportation Act. Most economists are favourable to deregulation and
freedom to enter, and the US experience indicated that cost reductions and
increased efficiency could follow. In 1995 an agreement was reached between
the US and Canada that provided full freedom for Canadian carriers to move
passengers to any US city, and freedom for US carriers to do likewise,
subject to a phase-in provision.

The National Energy Board regulates the development and transmission of oil
and natural gas. But earlier powers of the Board, involving the regulation
of product prices, were eliminated in 1986, and controls on oil exports were
also eliminated.

Agriculture remains a highly controlled area of the economy. Supply
`management', which is really supply restriction, and therefore `price
maintenance', characterizes grain, dairy, poultry and other products.
Management is primarily through provincial marketing boards.

\newhtmlpage

\subsection*{The role of the sharing economy}

The arrival of universal access to the internet has seen the emergence of
what is known as the \terminology{Sharing economy} throughout the world. 
This expression is used to descibe commercial
activities that, in the first place, are internet-based. Second, suppliers
in the sharing economy use resources in the market place that were initially
aimed at a different purpose. \textit{Airbnb} and \textit{Uber} are good
examples of companies in sectors of the economy where sharing is possible.
In Uber's case, the `ride-share' drivers initially purchased their vehicles
for private use, and subsequently redirected them to commercial use. Airbnb
is a communication corporation that enables the owners of spare home
capacity to sell the use of that capacity to short-term renters. With the
maturation of such corporations, the concept of `initial' and `secondary'
use becomes immaterial. 

\begin{DefBox}
	\textbf{Sharing economy}: involves enterprises that are internet based, and that use production resources that have use outside of the marketplace. 
\end{DefBox}

The importance of the sharing economy is that it provides an additional
source of competition to established suppliers, and therefore limits the
market power of the latter. At the same time, the emergence of the sharing
economy poses a new set of regulatory challenges: If traditional taxis are
required to purchase operating permits (medallions), and the ride-share
drivers do not require such permits, is there a reasonable degree of
competition in the market, and if not what is the appropriate solution?
Should the medallion requirement be abolished, or should ride-share drivers
be required to purchase one? In the case of \textit{Airbnb}, the suppliers
operate outside of the traditional `hotel' market. In general they do not
charge sales taxes or face any union labour agreements. What is the
appropriate response from governments? And how should the sharing economy be
taxed?

\newhtmlpage

\subsection*{Price regulation}

Regulating monopolistic sectors of the economy is one means of reducing
their market power. In Chapter~\ref{chap:imperfectcompetition} it was
proposed that indefinitely decreasing production costs in an industry means
that the industry might be considered as a `natural' monopoly: Higher output
can be produced at lower cost with fewer firms. Hence, a single supplier has
the potential to supply the market at a lower unit cost; unless, that is,
such a single supplier uses his monopoly power. To illustrate how the
consumer side may benefit from this production structure through regulation,
consider Figure~\ref{fig:deccostsupplier}. For simplicity suppose that
long-run marginal costs are constant and that average costs are downward
sloping due to an initial fixed cost. The profit-maximizing (monopoly)
output is where $MR=MC$ at $Q_m$ and is sold at the price $P_m$. This output
is inefficient because the willingness of buyers to pay for additional units
of output exceeds the additional cost. On this criterion the efficient
output is $Q^*$. But $LATC$ exceeds price at $Q^*$, and therefore it is not
feasible for a producer.

% Figure 14.4
\input{figures/ch14/ch14fig4}

\newhtmlpage

One solution is for the regulating body to set a price-quantity combination
of $P_r$, and $Q_r$, where price equals average cost and therefore generates
a normal rate of profit. This output level is still lower than the efficient
output level $Q^*$, but is more efficient than the profit-maximizing output 
$Q_m$. It is more efficient in the sense that it is closer to the efficient
output $Q^*$. A problem with such a strategy is that it may induce lax
management or overpayment to the factors of production: If producers are
allowed to charge an average-cost price, then there is a reduced incentive
for them to keep strict control of their costs in the absence of competition
in the marketplace.

A second solution to the declining average cost phenomenon is to implement
what is called a \terminology{two-part tariff}. This means that customers
pay an `entry fee' in order to be able to purchase the good. For example,
hydro or natural gas subscribers may pay a fixed charge per month for their
supply line and supply guarantee, and then pay an additional charge that
varies with quantity. In this way it is possible for the supplier to charge
a price per unit of output that is closer to marginal cost and still make a
profit, than under an average cost pricing formula. In terms of Figure~\ref{fig:deccostsupplier},
the total value of entry fees, or fixed components of
the pricing, would have to cover the difference between $MC$ and $LATC$
times the output supplied. In Figure~\ref{fig:deccostsupplier} this implies
that if the efficient output $Q^{*}$ is purchased at a price equal to
the $MC$ the producer loses the amount $(c-MC)$ on each unit sold. The
access fees would therefore have to cover at least this value.

Such a solution is appropriate when fixed costs are high and marginal costs
are low. This situation is particularly relevant in the modern market for
telecommunications: The cost to suppliers of marginal access to their
networks, whether it be for internet, phone or TV, is negligible compared to
the cost of maintaining the network and installing capacity.

\begin{DefBox}
	\textbf{Two-part tariff}: involves an access fee and a per unit of quantity fee.
\end{DefBox}

Finally, a word of caution: Nobel Laureate George Stigler has argued that
there is a danger of regulators becoming too close to the regulated, and
that the relationship can evolve to a point where the regulator may protect
the regulated firms. A likely such case was Canada's airline industry in the
seventies.
