\section{Business organization}\label{sec:ch7sec1}

Many different types of suppliers provide goods and services to the
marketplace. Some are small, some are large. But, whatever their size,
suppliers choose an organizational form that is appropriate for their
business: Aircraft and oil rigs are produced by large corporations; dental
services are provided by individual professionals or private partnerships.

The initial material of this chapter addresses organizational forms, their
goals and their operation. We then examine why individuals choose to invest
in firms, and propose that such investment provides individual investors
with a means both to earning a return on their savings and to managing the
risk associated with investing.

Understanding the way firms and capital markets function is crucial to
understanding our economic history and how different forms of social and
economic institutions interact. For example, seventeenth-century Amsterdam
had a thriving bourgeoisie, well-developed financial markets, and investors
with savings. This environment facilitated the channeling of investors'
funds to firms specializing in trade and nautical conquest. This tiny state
was then the source of some of the world's leading explorers and traders,
and it had colonies stretching to Indonesia. The result was economic growth
and prosperity.

In contrast, for much of the twentieth century, the Soviet Union dominated a
huge territory covering much of Asia and Europe. But capital markets were
non-existent, independent firms were stifled, and economic decline
ultimately ensued. Much of the enormous difference in the respective
patterns of economic development can be explained by the fact that one state
fostered firms, capital markets, and legal institutions, while the other did not. In terms of our
production possibility frontier: One set of institutional arrangements was
conducive to expanding the possibilities; the other was not. Sustainable new
businesses invariably require investors at an early point in the lifecycle
of the business. Accordingly, financial institutions that facilitate the
flow of savings and financial investment into new enterprises perform a
vital function in the economy.

Businesses, or firms, have several different forms. At the smallest scale, a
business takes the form of a \terminology{sole proprietor} or sole trader
who is the exclusive owner. A sole trader gets all of the revenues from the
firm and incurs all of the costs. Hence he may make profits or be personally
liable for the losses. In the latter case his business or even personal
assets may be confiscated to cover debts. Personal bankruptcy may result.

\begin{DefBox}
\textbf{Sole proprietor} is the single owner of a business.
\end{DefBox}

\newhtmlpage

If a business is to grow, \terminology{partners} may be required. Such
partners can inject money in exchange for a share of future profits. Firms
where trust is involved, such as legal or accounting firms, typically adopt
this structure. A firm is given credibility when customers see that partners
invest their own wealth in it.

\begin{DefBox}
\textbf{Partnership}: a business owned jointly by two or more individuals, who share in the profits and are jointly responsible for losses.
\end{DefBox}

If a business is to grow to a significant size it will generally need cash
and thus partners. Firms that provide legal services or dental services rely
primarily on human expertise, and therefore they need relatively little
physical capital. Hence their cash start-up needs tend to be modest. But
firms that produce aircraft need vast amounts of money to construct assembly
facilities; pharmaceuticals may need a billion dollars worth of research and
development to bring a new drug to the marketplace. Such businesses form
corporations -- also known as companies.

Large organizations have several inherent advantages over small
organizations when a high output level is required. Specialization in
particular tasks leads to increased efficiency for production workers. At
the same time, non-production workers can perform a multitude of different
tasks. If a large corporation decided to contract out every task involved in
bringing its product to market, the costs of such agreements could be
prohibitively high. In addition, synergies can arise from teamwork. New
ideas and better work flow are more likely to materialize when individuals
work in close proximity than when working as isolated units, no matter how
efficient they may be individually. A key aspect of such large organizations
is that \textit{they have a legal identity separate from the managers and
owners}.

\begin{DefBox}
\textbf{Corporation or company} is an organization with a legal identity separate from its owners that produces and trades.
\end{DefBox}

\newhtmlpage

The owners of a corporation are known as its \terminology{shareholders}, and
their object is usually to make profits. There also exist non-profit
corporations whose objective may be philanthropic. Since our focus is upon
markets, we will generally assume that profits form the objective of a
typical corporation. The profits that accrue to a corporation may be paid to
the shareholders in the form of a \terminology{dividend}, or retained in the
corporation for future use. When large profits (or losses) accrue the value
of the corporation increases (or decreases), and this is reflected in the
value of each share of the company. If the value of each share in the
company increases (decreases) there is a \terminology{capital gain (loss)}
to the owners of the shares -- the shareholders. In any given year
shareholders can receive a dividend and also obtain a capital gain (or
loss). The sum of the dividend and capital gain represents the return to
owning their corporate stock in that year. When this sum is adjusted for
inflation it is termed the \terminology{real return on corporate stock}

\begin{DefBox}
\textbf{Shareholders} invest in corporations and therefore are the owners. 

\textbf{Dividends} are payments made from after-tax profits to company shareholders.

\textbf{Capital gains (losses)} arise from the ownership of a corporation when an individual sells a share at a price higher (lower) than when the share was purchased.

\textbf{Real return to corporate stock} is the inflation-adjusted sum of dividends and capital gain (or loss).
\end{DefBox}

\newhtmlpage

A key difference between a company and a partnership is that a company
involves \terminology{limited liability}, whereas a partnership does not.
Limited liability means that the liability of the company is limited to the
value of the company's assets. Shareholders cannot be further liable for any
wrongdoing on the part of the company. Accordingly, partnerships and sole
traders frequently insure themselves and their operations. For example, all
specialist doctors carry malpractice insurance.

\begin{DefBox}
\textbf{Limited liability} means that the liability of the company is limited to the value of the company's assets.
\end{DefBox}

Corporations use capital, labour, and human expertise to produce a good, to
supply a service, or to act as an intermediary. Corporations are required to
produce an annual income statement that accurately describes the operation
of the firm. An example is given in Table~\ref{table:BMO2008}.

%Table 7.1
\begin{Table}{caption={Bank of Montreal income statement, 2008 \label{table:BMO2008}}}
\begin{tabu} to 30em {|X[5,l]X[0.25,c]X[1,r]|} \hline 
	\rowcolor{rowcolour}	Net income (after tax) from operations	&	\$	&	1,978m 		\\
	Preferred dividends						&	\$	&	73m 			\\
	\rowcolor{rowcolour}	Income available to shareholders		&	\$	&	1,905m 		\\
	Number of common stocks					&	\$	&	502m 			\\
	\rowcolor{rowcolour}	Earnings per common share				&	\$	&	3.79 			\\
	Dividends per common share				&	\$	&	2.80 			\\
	\rowcolor{rowcolour}											&		&					\\
	Market capitalization  (June 2008)		&	\$	&	23.7b			\\
	\rowcolor{rowcolour}	Stock price  (June 2008)				&	\$	&	48.0    		\\ \hline 
\end{tabu}
\end{Table}

\newhtmlpage

``Net income from operations'' represents
after-tax profits. Of this amount, \$73 million was paid to holders of a
special class of shares, and \$1.905 million was available for either
reinvesting in the company or payment of dividends. There were 502 million
shares outstanding in the company at the end of 2008. This means that if the
\$1.905 million were allocated over all such shares, \$3.79 could go to each
one. But corporations normally retain a substantial part of such earnings in
order to finance investment for coming years. This component is called the %
\terminology{retained earnings}. The remainder is distributed to the
owners/shareholders in the form of dividends. In this case, of the \$3.79
earnings per share, \$2.80 was distributed to shareholders and the remainder
was retained inside the company. The final two entries in the table indicate
that the total value of the company's stock was \$23.7 billion, and each
unit of stock was valued at approximately \$48 in June 2008.

Such information is publicly available for a vast number of corporations at
the `finance' section of major search engines such as \textit{Google} or \textit{Yahoo}.

\begin{DefBox}
\textbf{Retained earnings} are the profits retained by a company for reinvestment and not distributed as dividends.
\end{DefBox}

In Canada, the corporate sector as a whole tends to hold on to more than
half of after-tax profits in the form of retained earnings. However there
exists considerable variety in the behaviour of corporations, and each firm
tends to have a known and established pattern of how it allocates its profits
between dividends and retained earnings. In the Table~\ref{table:BMO2008}
example, three quarters of profits were distributed; yet some corporations
have a no-dividend policy. In these latter cases the benefit to investing in
a firm must come in the form of capital gain to the owners of the shares.