\newpage
\section{The time horizon and inflation}\label{sec:ch4sec3}

The price elasticity of demand is frequently lower in the short run than in
the long run. For example, a rise in the price of home heating oil may
ultimately induce consumers to switch to natural gas or electricity, but
such a transition may require a considerable amount of time. Time is
required for decision-making and investment in new heating equipment. A
further example is the elasticity of demand for tobacco. Some adults who
smoke may be seriously dependent and find quitting almost impossible. But if
young smokers, who are not yet addicted, decide not to start on account of
the higher price, then over a long period of time the percentage of the
population that smokes will decline. The full impact may take decades!
Accordingly when we talk of the short run and the long run, there is no simple rule for
defining how long the long run actually is in terms of months or years. In
some cases, adjustment may be complete in weeks, in other cases years.

In Chapter~\ref{chap:tmd} we distinguished between real and nominal
variables. The former adjust for inflation; the latter do not. Suppose all
nominal variables double in value: Every good and service costs twice as
much, wage rates double, dividends and rent double, etc. This implies that
whatever bundle of goods was previously affordable is still affordable.
Nothing has really changed. Demand behaviour is unaltered by this doubling
of all prices and all incomes.

How do we reconcile this with the idea that own-price elasticities measure
changes in quantity demanded as prices change? Keep in mind that
elasticities measure the impact of changing one variable alone, \textit{%
holding constant all of the others}. But when all variables are changing
simultaneously, it is incorrect to think that the impact on quantity of one
price or income change is a true measure of responsiveness or elasticity.
The \textit{price changes that go into measuring elasticities are therefore
changes in prices relative to inflation}.