\section{The income elasticity of demand}\label{sec:ch4sec5}

In Chapter~\ref{chap:classical} we stated that higher incomes tend to
increase the quantity demanded at any price. To measure the responsiveness
of demand to income changes, a unit-free measure exists: The income
elasticity of demand. The \terminology{income elasticity of demand} is the
percentage change in quantity demanded divided by a percentage change in
income.

\begin{DefBox}
The \textbf{income elasticity of demand} is the percentage change in quantity demanded divided by a percentage change in income.
\end{DefBox}

Let us use the Greek letter eta, $\eta$, to define the income elasticity of
demand and $I$ to denote income. Then,

\begin{equation*}
\eta_{d}=\frac{\text{percentage change in quantity demanded}}{\text{%
percentage change in income}}=\frac{\%\Delta Q}{\%\Delta I} 
\end{equation*}

As an example, if monthly income increases by 10 percent, and the quantity
of magazines purchased increases by 15 percent, then the income elasticity
of demand for magazines is 1.5 in value $(=15\%/10\%)$. The income
elasticity is generally positive, but not always -- let us see why.

\newhtmlpage

\subsection*{Normal, inferior, necessary, and luxury goods}

The income elasticity of demand, in diagrammatic terms, is a percentage
measure of how far the demand curve shifts in response to a change in
income. Figure~\ref{fig:incomeelastdemandshift} shows two possible shifts.
Suppose the demand curve is initially the one defined by $D$, and then
income increases. In this example the supply curve is horizontal at the
price $P_0$. If the demand curve shifts to $D_1$ as a result, the
change in quantity demanded at the existing price is $(Q_1-Q_0)$. However,
if instead the demand curve shifts to $D_2$, that shift denotes a larger
change in quantity $(Q_2-Q_0)$. Since the shift in demand denoted by $D_2$
exceeds the shift to $D_1$, the $D_2$ shift is more responsive to income,
and therefore implies a higher income elasticity.

% Figure 4.6	fig:incomeelastdemandshift
\input{figures/ch4/ch4fig6}

\newhtmlpage

In this example, the good is a \textit{normal good}, as defined in Chapter~%
\ref{chap:classical}, because the demand for it increases in response to
income increases. If the demand curve were to shift back to the left in
response to an increase in income, then the income elasticity would be
negative. In such cases the goods or services are \textit{inferior}, as
defined in Chapter~\ref{chap:classical}.

Finally, we distinguish between luxuries and necessities. A %
\terminology{luxury} good or service is one whose income elasticity equals
or exceeds unity. A \terminology{necessity} is one whose income elasticity
is greater than zero but less than unity. If quantity demanded is so
responsive to an income increase that the percentage increase in quantity
demanded exceeds the percentage increase in income, then the elasticity
value is in excess of 1, and the good or service is called a luxury. In
contrast, if the percentage change in quantity demanded is less than the
percentage increase in income, the value is less than unity, and we call the
good or service a necessity.

\begin{DefBox}
A \textbf{luxury} good or service is one whose income elasticity equals or exceeds unity.
	
A \textbf{necessity} is one whose income elasticity is greater than zero and less than unity.
\end{DefBox}

Luxuries and necessities can also be defined in terms of their share of a
typical budget. An income elasticity greater than unity means that the share
of an individual's budget being allocated to the product is increasing. In
contrast, if the elasticity is less than unity, the budget share is falling.
This makes intuitive sense---luxury cars are luxury goods by this definition
because they take up a larger share of the incomes of the rich than the
non-rich.

\newhtmlpage

\terminology{Inferior goods} are those for which there exist higher-quality,
more expensive, substitutes. For example, lower-income households tend to
satisfy their travel needs by using public transit. As income rises,
households normally reduce their reliance on public transit in favour of
automobile use (despite the congestion and environmental impacts). Inferior
goods, therefore, have a negative income elasticity: In the income
elasticity equation definition, the numerator has a sign opposite to that of
the denominator.

\begin{DefBox}
\textbf{Inferior goods} have negative income elasticity.
\end{DefBox}

Empirical research indicates that goods like food and fuel have income
elasticities less than 1; durable goods and services have elasticities
slightly greater than 1; leisure goods and foreign holidays have
elasticities very much greater than 1.

\textit{Income elasticities are useful in forecasting the demand for
particular services and goods in a growing economy}. Suppose real income is
forecast to grow by 15\% over the next five years. If we know that the
income elasticity of demand for smart phones is 2.0, we could estimate the
anticipated growth in demand by using the income elasticity formula: Since
in this case $\eta=2.0$ and $\%\Delta I=15$ it follows that $2.0=\%\Delta
Q/15\%$. Therefore the predicted demand change must be 30\%.