%\newpage
\section{Elasticity of supply}\label{sec:ch4sec6}

Now that we have developed the various dimensions of elasticity on the
demand side, the analysis of elasticities on the supply side is
straightforward. The \terminology{elasticity of supply} measures the
responsiveness of the quantity supplied to a change in the price.

\begin{DefBox}
The \textbf{elasticity of supply} measures the responsiveness of quantity supplied to a change in the price.
\end{DefBox}

\begin{equation*}
\varepsilon_{s}=\frac{\text{percentage change in quantity supplied}}{\text{%
percentage change in price}}=\frac{\%\Delta Q}{\%\Delta P} 
\end{equation*}

The subscript $s$ denotes supply. This is exactly the same formula as for
the demand curve, except that the quantities now come from a supply curve.
Furthermore, and in contrast to the demand elasticity, the supply elasticity
is generally a positive value because of the positive relationship between
price and quantity supplied. The more elastic, or the more responsive, is
supply to a given price change, the larger will be the elasticity value. In
diagrammatic terms, this means that ``flatter'' supply curves have a greater
elasticity than more ``vertical'' curves at a given price and quantity
combination. Numerically the flatter curve has a larger value than the more
vertical supply -- try drawing a supply diagram similar to Figure~\ref{fig:limitingcasepriceelasticity}.
Technically, a completely vertical supply
curve has a zero elasticity and a horizontal supply curve has an infinite
elasticity -- just as in the demand cases.

As always we keep in mind the danger of interpreting too much about the
value of this elasticity from looking at the visual profiles of supply
curves.
