\section{Human capital}\label{sec:ch13sec1}

\terminology{Human capital}, HK, is the stock of knowledge and ability
accumulated by a worker that determines future productivity and earnings. It
depends on many different attributes -- education, experience, intelligence,
interpersonal skills etc. While individuals look upon human capital as a
determinant of their own earnings, it also impacts the productivity of the
economy at large, and is therefore a vital force in determining long-run
growth. Canada has been investing heavily in human
capital in recent decades, and this suggests that future productivity and
earnings will benefit accordingly.

\begin{DefBox}
	\textbf{Human capital} is the stock of knowledge and ability accumulated by a worker that determines future productivity and earning.
\end{DefBox}

Several features of Canada's recent human capital accumulation are
noteworthy. First, Canada's enrolment rate in post-secondary education now
exceeds the US rate, and that of virtually every economy in the world.
Second is the fact that the number of women in third-level institutions
exceeds the number of men. Almost 60\% of university students are women.
Third, international testing of high-school students sees Canadian students
performing well, which indicates that the quality of the Canadian
educational system appears to be high. These are positive aspects of a
system that frequently comes under criticism. Nonetheless, the distribution
of income that emerges from market forces in Canada has become more unequal.

Let us now try to understand individuals' motivation for embarking on the
accumulation of human capital, and in the process see why different groups
earn different amounts. We start by analyzing the role of education and then
turn to on-the-job training.