\section{Education as signalling}\label{sec:ch13sec4}

An alternative view of education springs from the theory of signalling. This
is a provocative theory that proposes education may be worthwhile, even if
it generates little additional skills. The theory recognizes that
individuals possess different abilities. However, firms cannot easily
recognize the more productive workers without actually hiring them and
finding out \textit{ex-post} -- sometimes a costly process. %
\terminology{Signalling theory} says that, in pursuing more education,
people who know they are more capable thereby send a signal to potential employers
that they are the more capable workers. Education therefore %
\terminology{screens} out the low-productivity workers from the
high-productivity (more educated) workers. Firms pay more to the latter,
because firms know that the high-ability workers are those with the
additional education.

\begin{DefBox}
	\textbf{Signalling} is the decision to undertake an action in order to reveal information.
	
	\textbf{Screening} is the process of obtaining information by observing differences in behaviour.
\end{DefBox}

To be effective, the process must separate the two types. Why don't
lower-ability workers go to university and pretend they are of the
high-ability type? Primarily because that strategy could backfire: Such
individuals are less likely to succeed at school and there are costs
associated with school in the form of school fees, books and foregone
earnings. While they may have lower innate skills, they are likely smart
enough to recognize a bad bet.

On balance economists believe that further education does indeed add to
productivity, although there may be an element of screening present: An
engineering degree (we should hope) increases an individual's understanding
of mechanical forces so that she can design a bridge that will not collapse,
in addition to telling a potential employer that the student is smart!

Finally, it should be evident that if education raises productivity, it is
also good for society and the economy at large.