\section{Wealth and capitalism}\label{sec:ch13sec8}

In an insightful and popular study of capital accumulation, from both a
historical and contemporary perspective, Thomas Piketty draws our attention
to the enormous inequality in the distribution of wealth and explores what
the future may hold in his book \textit{Capital in the Twenty-First Century}.

The distribution of wealth is universally more unequal than the distribution
of incomes or earnings. Gini coefficients in the neighbourhood of 0.8 are
commonplace in developed economies. In terms of shares of the wealth pie,
such a magnitude may imply that the top 1\% of wealth holders own one third
of all of an economy's wealth, that the top decile may own two-thirds of all
wealth, and that the remaining one third is held by the `bottom 90\%'. And
within this bottom 90\%, virtually all of the remaining wealth is held by
the 40\% of the population below the top decile, leaving only a few percent
of all wealth to the bottom 50\% of the population.

% Figure 13.5
\input{figures/ch13/ch13fig5}

\newhtmlpage

While such an unequal holding pattern may appear shockingly unjust, Piketty
informs us that current wealth inequality is not as great in most economies
as it was about 1900. Figure~\ref{fig:wealthshareuseurope} (Piketty 10.6)
above is borrowed from his web site. Wealth was more unequally distributed
in Old World Europe than New World America a century ago, but this
relativity has since been reversed. A great transformation in the wealth
holding pattern of societies in the twentieth century took the form of the
emergence of a `patrimonial middle class', by which he means the emergence
of substantial wealth holdings on the part of that 40\% of the population
below the top decile. This development is noteworthy, but Piketty warns that
the top percentiles of the wealth distribution may be on their way to
controlling a share of all wealth close to their share in the early years of
the twentieth century. He illustrates that two elements are critical to this
prediction; first is the rate of growth in the economy relative to the
return on capital, the second is inheritances.

To illustrate: Imagine an economy with very low growth, and where the owners
of capital obtain an annual return of say 5\%. If the owners merely maintain
their capital intact and consume the remainder of this 5\%, then the pattern
of wealth holding will continue to be stable. However, if the holders of
wealth can reinvest from their return an amount more than is necessary to
replace depreciation then their wealth will grow. And if labour income in
this economy is almost static on account of low overall growth, then wealth
holders will secure a larger share of the economic pie. In contrast, if
economic growth is significant, then labour income may grow in line with
income from capital and inequality may remain stable. This summarizes
Piketty's famous $(r-g)$ law -- inequality depends upon the difference
between the return on wealth and the growth rate of the economy. This
potential for an ever-expanding degree of inequality is magnified when the
stock of capital in the economy is large.

Consider now the role of inheritances. That is to say, do individuals leave
large or small inheritances when they die, and how concentrated are such
inheritances? If individual wealth accumulation patterns are generated by a
desire to save for retirement and old age -- during which time individuals
decumulate by spending their assets -- such motivation should result in small
bequests being left to following generations. In contrast, if individuals
who are in a position to do so save and accumulate, not just for their old
age, but because they have dynastic preferences, or if they take pleasure
simply from the ownership of wealth, or even if they are very cautious about
running down their wealth in their old age, then we should see
substantial inheritances passed on to the sons and daughters of these
individuals, thereby perpetuating, and perhaps exacerbating, the inequality
of wealth holding in the economy.

Piketty shows that in fact individuals who save substantial amounts tend to
leave large bequests; that is they do not save purely for life-cycle
motives. In modern economies the annual amount of bequests and gifts from
parents to children falls in the range of 10\% to 15\% of annual GDP. This
may grow in future decades, and since wealth is highly concentrated, these
bequests in turn are concentrated among a small number of the following
generation -- inequality is transmitted from one generation to the next.

\newhtmlpage

As a final observation, if we consider the distribution of income and wealth
together, particularly at the very top end, we can see readily that a
growing concentration of \textit{income} among the top 1\% should ultimately
translate itself into greater \textit{wealth} inequality. This is because
top earners can save more easily than lower earners. To compound matters, if
individuals who inherit wealth also tend to inherit more human capital from
their parents than others, the concentration of income and wealth may become
yet stronger.

The study of distributional issues in economics has probably received too
little attention in the modern era. Yet it is vitally important both in
terms of the well-being of the individuals who constitute an economy and in
terms of adherence to social norms. Given that utility declines with
additions to income and wealth, transfers from those at the top to those at
the bottom have the potential to increase total utility in the economy.
Furthermore, an economy in which justice is seen to prevail---in the form of
avoiding excessive inequality---is more likely to achieve a higher degree of
social coherence than one where inequality is large.