\section{Monopolies}\label{sec:ch10sec1}

In analyzing perfect competition we emphasized the difference between the
industry and the individual supplier. The individual supplier is an
atomistic unit with no market power. In contrast, a monopolist has a great
deal of market power, for the simple reason that a \terminology{monopolist}
is the sole supplier of a particular product and so really \textit{is} the
industry. The word monopoly, comes from the Greek words \textit{monos},
meaning one, and \textit{polein} meaning to sell. When there is just a
single seller, our analysis need not distinguish between the industry and
the individual firm. They are the same on the supply side.

Furthermore, the distinction between long run and short run is blurred,
because a monopoly that continues to survive as a monopoly obviously sees no
entry or exit. This is not to pretend that monopolized sectors of the
economy do not `die' and reincarnate themselves in some other more
competitive form; they do. But if a monopoly continues to exist, by
definition there is no entry or exit in the industry.

\begin{DefBox}
	A \textbf{monopolist} is the sole supplier of an industry's output, and therefore the industry and the firm are one and the same.
\end{DefBox}

Economists' thinking on monopoly has evolved considerably in recent decades.
In part this is due to the impact of the opening up of national borders, in
part due to the impact of technological change. The world is now more open
and more subject to change, and so too are the circumstances governing
monopolies. There are three main reasons for the existence and continuance
of monopolies: Scale economies, national policy and successful prevention of
entry.

\newhtmlpage

\subsection*{Natural monopolies}

Traditionally, monopolies were viewed as being `natural' in some sectors of
the economy. This means that scale economies define some industries'
production and cost structures up to very high output levels, and that the
whole market might be supplied at least cost by a single firm.

Consider the situation depicted in Figure~\ref{fig:naturalmonopolist}. The
long-run $ATC$ curve declines indefinitely. There is no output level where
average costs begin to increase. Imagine now having several firms, each
producing with a plant size corresponding to the short-run average cost
curve $ATC_{1}$, or alternatively a single larger firm using a plant size
denoted by $ATC_{2}$. The small firms in this case cannot compete with the
larger firm because the larger firm has lower production costs and can
undercut the smaller firms, and \textit{supply the complete market} in the
process. Such a scenario is termed a \terminology{natural monopoly}.

% Figure 10.1
\input{figures/ch10/ch10fig1}

\newhtmlpage

\begin{DefBox}
	\textbf{Natural monopoly}: one where the $ATC$ of producing any output declines with the scale of operation.
\end{DefBox}

Electricity distribution in most of Canada's provinces is in the hands of a
single supplier -- \textit{Hydro Quebec} or \textit{Hydro One} in Ontario, for example. These
distributors are natural monopolies in the sense described above: Unit
distribution costs decline with size. In contrast, electricity production is
not `naturally' a monopoly. Other suppliers were once thought of as
`natural' monopolies also, but are no longer. \textit{Bell
	Canada} was considered to be a natural monopoly in the era of land lines: It
would not make economic sense to run several sets of phone lines to every
residence. But that was before the arrival of cell phones. \textit{Canada
	Post} was also thought to be a natural monopoly, until the advent of 
\textit{FEDEX}, \textit{UPS} and other couriers proved otherwise. Thus there is no
guarantee that even a self-perceived natural monopoly can continue to hold
that status indefinitely. Invention can compete that status away.

In reality there are very few \textit{pure} monopolies. \textit{Facebook}, 
\textit{Microsoft} and \textit{Google} may be extraordinarily dominant in their
markets, but they are not the only suppliers of the services or products
that they offer. There exist other products that are similar.

\subsection*{National Policy}

A second reason for monopolies is national policy. Some governments are, or
once were, proud to have a `national carrier' in the airline industry -- \textit{Air
Canada} in Canada or \textit{British Airways} in the UK. The mail service was viewed
as a symbol of nationhood in Canada and the US: \textit{Canada Post} and the 
\textit{US Postal} system are national emblems that have historic
significance. They were vehicles for integrating the provinces or states at
various points in the federal lives of these countries.

The down side of such nationalist policies is that they can be costly to the
taxpayer. Industries that are not subject to competition can become fat and
uncompetitive: Managers have insufficient incentives to curtail costs;
unions realize the government is committed to sustain the monopoly and push
for higher wages than under a more competitive structure.

\newhtmlpage

\subsection*{Maintaining barriers to entry}

A third reason that monopolistic companies continue to survive is that they
are successful in preventing the entry of new firms and products. 
\textit{Patents and copyrights} are one vehicle for preserving the sole-supplier
role, and are certainly necessary to encourage firms to undertake the
research and development (R\&D) for new products.

Many corporations produce products that require a large up-front investment;
this might be in the form of research and development, or the construction
of costly production facilities. For example, \textit{Boeing} or \textit{%
	Airbus} incurs billions of dollars in developing new aircraft;
pharmaceuticals may have to invest a billion dollars to develop a new drug.
However, once such an investment is complete, the cost of producing each
unit of output may be very low. Such a phenomenon is displayed in Figure~\ref{fig:fixedconstmarginalcost}.
In this case the average cost for a small
number of units produced is high, but once the fixed cost is spread over an
ever larger output, the average cost declines prapidly, and in the limit
approaches the marginal cost. These production structures are common in
today's global economy, and they give rise to markets characterized either
by a single supplier or a small number of suppliers.

% Figure 10.2
\input{figures/ch10/ch10fig2}

\newhtmlpage

This figure is useful in understanding the role of patents. Suppose that 
\textit{Pharma A} spends one billion dollars in developing a new drug and
has constant unit production costs thereafter, while \textit{Pharma B}
avoids research and development and simply imitates \textit{Pharma A}'s
product. Clearly \textit{Pharma B} would have a $LATC$ equal to its $LMC$,
and would be able to undercut the initial developer of the drug. Such an
outcome would discourage investment in new products and the economy at large
would suffer as a consequence. Economies would be worse off if protection is
not provided to the developers of new products, because if such protection
is not offered potential developers will not have the incentive to incur the
up-front investment required.

While copyright and patent protection is legal, \textit{predatory pricing}
is an illegal form of entry barrier, and we explore it more fully in Chapter~%
\ref{chap:government}. An example would be where an existing firm that sells
nationally may deliberately undercut the price of a small local entrant
to the industry. Airlines with a national scope are frequently accused of
posting low fares on flights in regional markets that a new carrier is
trying to enter.

\textit{Political lobbying} is another means of maintaining monopolistic
power. For example, the \textit{Canadian Wheat Board} had fought
successfully for decades to prevent independent farmers from marketing
wheat. This Board lost its monopoly status in August 2012, when the
government of the day decided it was not beneficial to consumers or farmers
in general. Numerous `supply management' policies are in operation all
across Canada. Agriculture is protected by production quotas. All maple
syrup in Quebec must be marketed through a single monopoly supplier.

\textit{Critical networks} also form a type of barrier, though not always a
monopoly. The compression software \textit{WinZip} benefits from the fact
that it is used almost universally. There are many free compression programs
available on the internet. But \textit{WinZip} can charge for its product
because virtually everybody uses it. Free network products that no one uses
are worth very little, even if they come at zero cost.