\section{Long-run choices}\label{sec:ch10sec3}

Consider next the impact of a shift in demand upon the profit maximizing
choice of this firm. A rightward shift in demand in Figure~\ref{fig:monopolyeq}
also yields a new $MR$ curve. The firm therefore chooses a
new level of output, using the same profit maximizing rule: Set $MC=MR$.
This output will be greater than the previous output, but again the price
must be on an elastic portion of the new demand curve. If operating with the
same plant size, the $MC$ and $ATC$ curves do not change and the new profit
per unit is again read from the $ATC$ curve.

By this stage the curious student will have asked: ``What happens to plant
size in the long run?'' For example, is the monopolist in Figure~\ref{fig:monopolyeq}
using the most appropriate plant size in the first place?
Even if she is, should the monopolist consider adopting an expanded plant
size in response to the shift in demand?

The answer is: In the long run the monopolist is free to choose whatever
plant size is best. Her initial plant size might have been optimal for the
demand she faced, but if it was, it is unlikely to be optimal for the larger
scale of production associated with the demand shift. Accordingly, with the
new demand curve, she must consider how much profit she could make using
different plant sizes.

\newhtmlpage

% Figure 10.8
\input{figures/ch10/ch10fig8}

To illustrate one possibility, we will think of this firm as having constant
returns to scale at all output ranges, as displayed in Figure~\ref{fig:monopolyplantsize}.
(Our reasoning carries through if the $LAC$ slopes
downwards; the graph just becomes a little more complex.) The key
characteristic of constant returns to scale is that a doubling of inputs
leads to a doubling of output. Therefore, if the per-unit cost of inputs is
fixed, a doubling of inputs (and therefore output) leads exactly to a
doubling of costs. This implies that, when the firm varies its plant size 
\textit{and} its labour use, the cost of producing each additional unit must
be constant. The long-run marginal cost $LMC$ is therefore constant and
equals the $ATC$ in the long run.

\newhtmlpage
%\newpage

Figure~\ref{fig:plantsizeLR} describes the market for this good. The optimal
output and price are determined in the usual manner: Set $MC=MR$. If the
monopolist has plant size corresponding to $ATC_1$, the optimal output is 
$Q_1$ and should be sold at the price $P_1$.The key issue now is: Given the
demand conditions, could the monopolist make more profit by choosing a plant
size that differs from the one corresponding to $ATC_1$?

% Figure 10.9
\input{figures/ch10/ch10fig9}

In this instance the answer is a clear `yes'. Her $LMC$ curve is horizontal
and so, by increasing output from $Q_1$ to $Q_2$ she earns a profit on each
additional unit in that range, because the $MR$ curve lies above the $LMC$
curve. In order to produce the output level $Q_2$ at least cost she must
choose a plant size corresponding to $AC_2$.
