\section{Trade in our daily lives}\label{sec:ch15sec1}

Virtually every economy in the modern world trades with other economies --
they are what we call `open' economies. Evidence of such openness is
everywhere evident in our daily life. The world eats Canadian wheat; China
exports manufactured goods to almost anywhere we can think of; and Canadians
take their holidays in Florida.

As consumers we value the choice and variety of products that trade offers.
We benefit from lower prices than would prevail in a world of protectionism.
At the same time there is a constant chorus of voices calling for protection
from international competition: Manufacturers are threatened by production
in Asia; dairy farmers cry out against the imports of poultry, beef, and
dairy products; even the service sector is concerned about offshore
competition from call centres and designers. In this world of competing
views it is vital to understand how trade has the potential to improve the
well-being of economies.

This chapter examines the theory of international trade, trade flows, and
trade policy: Who trades with whom, in what commodities, and why. In
general, countries trade with one another because they can buy foreign
products at a lower price than it costs to make them at home. International
trade reflects specialization and exchange, which in turn improve living
standards. It is cost differences between countries rather than
technological differences that drive trade: In principle, Canada could
supply Toronto with olives and oranges grown in Nunavut greenhouses, but it
makes more sense to import them from Greece, Florida or Mexico.

Trade between Canada and other countries differs from trade between
provinces. By definition, international trade involves jumping a border,
whereas most trade within Canada does not. Internal borders are present in
some instances -- for example when it comes to recognizing professional
qualifications acquired out-of-province. In the second instance,
international trade may involve different currencies. When Canadians trade
with Europeans the trade is accompanied by financial transactions involving
Canadian dollars and Euros. A Canadian buyer of French wine pays in Canadian
dollars, but the French vineyard worker is paid in euros. Exchange rates are
one factor in determining national competitiveness in international markets.
Evidently, not every international trade requires currency trades at the
same time -- most members of the European Union use the Euro. Indeed a
common currency was seen as a means of facilitating trade between member
nations of the EU, and thus a means of integrating the constituent economies
more effectively.