\section{The gains from trade: Comparative advantage }\label{sec:ch15sec3}

In the opening chapter of this text we emphasized the importance of
opportunity cost and differing efficiencies in the production process as a
means of generating benefits to individuals through trade in the
marketplace. The simple example we developed illustrated that, where
individuals differ in their efficiency levels, benefits can accrue to each
individual as a result of specializing and trading. In that example it was
assumed that individual A had an absolute advantage in producing one product
and that individual Z had an absolute advantage in producing the second
good. This set-up could equally well be applied to two economies that have
different efficiencies and are considering trade, with the objective of
increasing their consumption possibilities. Technically, we could replace
Amanda and Zoe with Argentina and Zambia, and nothing in the analysis would
have to change in order to illustrate that consumption gains could be
attained by both Argentina and Zambia as a result of specialization and
trade.

\textit{Remember}: The opportunity cost of a good is the quantity of another
good or service given up in order to have one more unit of the good in
question.

So, let us now consider two \underbar{economies} with differing production
capabilities, as illustrated in Figures~\ref{fig:compadvprod} and~\ref{fig:compadvcons}.
In this instance it is assumed that one economy has an
absolute advantage in both goods, but the degree of that advantage is
greater in one good than the other. In international trade language, there
exists a comparative advantage as well as an absolute advantage. It is
frequently a surprise to students that this situation has the capacity to
yield consumption advantages to each economy, even though one is absolutely
more efficient in producing both of the goods. This is termed the %
\terminology{principle of comparative advantage}, and it states that even if
one country has an absolute advantage in producing both goods, gains to
specialization and trade still materialize, provided the opportunity cost of
producing the goods differs between economies. This is a remarkable result,
and much less intuitive than the principle of absolute advantage. We explore
it with the help of the example developed in Figures~\ref{fig:compadvprod}
and \ref{fig:compadvcons}.

\begin{DefBox}
	\textbf{Principle of comparative advantage} states that even if one country has an absolute advantage in producing both goods, gains to specialization and trade still materialize, provided the opportunity cost of producing the goods differs between economies.
\end{DefBox}

\newhtmlpage

The two economies considered are the US and Canada. Their production
possibilities are defined by the $PPF$s in Figure~\ref{fig:compadvprod}.
Canada can produce 5 units of $V$ or 35 units of $F$, or any combination
defined by the line joining these points. With the same resources the US 
can produce 8$V$ or 40$F$, or any combination defined by its $PPF$\footnote{%
	Note that we are considering the $PPF$s to be straight lines rather than
	concave shapes. The result we illustrate here carries over to that case
	also, but it is simpler to illustrate with the linear $PPF$s.}. Canadians
and Americans consume a combination of the goods defined by some point on
their respective $PPF$s. The opportunity cost of a unit of $V$ in Canada is 7$F$ 
(the slope of Canada's $PPF$ is $5/35=1/7$). In the US the opportunity
cost of one unit of $V$ is 5$F$ (slope is $8/40=1/5$). In this set-up the US
is more efficient in producing $V$ than $F$ relative to Canada, as reflected
by the opportunity costs. Hence we say that the \textit{US has a comparative
	advantage in the production of $V$} \underbar{and} that \textit{Canada has
	therefore a comparative advantage in producing $F$}.

% Figure 15.1
\input{figures/ch15/ch15fig1}

\newhtmlpage

Prior to trade each economy is producing all of the goods it consumes. This
no-trade state is termed \terminology{autarky}.

\begin{DefBox}
	\textbf{Autarky} denotes the no-trade situation.
\end{DefBox}

\newhtmlpage

\subsection*{The gains from trade}

We now permit each economy to specialize in producing where it has a
comparative advantage. So Canada specializes completely by producing 35$F$
and the US produces 8$V$. Having done this the economies must now agree on
the terms of trade. The \terminology{terms of trade} define the rate at
which the two goods will trade post-specialization. Let us suppose that a
bargaining process leads to agreement that one unit of $V$ will trade for
six units of $F$. Such a trading rate, one that lies between the opportunity
costs of each economy, benefits both economies. This exchange rate lies 
between Canada's opportunity cost of $1:7$ and the US opportunity cost of $1:5$. 
By specializing in $F$, Canada can now obtain an additional unit of $V$ by sacrificing six units of 
$F$, whereas pre-trade it had to sacrifice seven units of $F$ for a unit of $V$. 
Technically, by specializing in $F$ and trading at a rate of $1:6$ Canada's
consumption possibilities have expanded and are given by the consumption
possibility frontier ($CPF$) illustrated in Figure~\ref{fig:compadvcons}.
The \terminology{consumption possibility frontier} defines what an economy
can consume after production specialization and trade.

% Figure 15.2
\input{figures/ch15/ch15fig2}

\newhtmlpage

The US also experiences an improved set of consumption possibilities. By
specializing in $V$ and trading at a rate of $1:6$ its $CPF$ lies outside its $%
PPF$ and this enables it to consume more than in the pre-specialization
state, where its $CPF$ was defined by its $PPF$.

Evidently, the US and Canada CPFs are parallel since they trade with each
other at the same rate: If Canada exports six units of $F$ for every unit of 
$V$ that it imports from the US, then the US must import the same six units
of $F$ for each unit of $V$ it exports to Canada. The remarkable outcome
here is that, even though one economy is more efficient in producing each
good, specialization still leads to gains for both economies. The gain is
illustrated by the fact that each economy's consumption possibilities lie
outside of its production possibilities\footnote{%
To illustrate the gains numerically, let Canada import 3$V$ from the US in
return for exporting 18$F$. Note that this is a trading rate of $1:6$. Hence,
Canada consumes 3$V$ and 17$F$ (Canada produced 35$F$ and exported 18$F$,
leaving it with 17$F$). It follows that the US consumes 5$V$, having
exported 3$V$ of the 8$V$ it produced, and obtained in return 18$F$ in
imports. The new consumption bundles are illustrated in the figure: $(17,3)
$ for Canada and $(18,5)$ for the US. These consumption bundles clearly
represent an improvement over the autarky situation. For example, if Canada
had wished to consume 3$V$ pre-trade, it would only have been able to
consume 14$F$, whereas with trade it can consume 17$F$.}.

\begin{DefBox}
	\textbf{Terms of trade} define the rate at which the goods trade internationally.
	
	\textbf{Consumption possibility frontier} defines what an economy can consume after production specialization and trade.
\end{DefBox}
	
\newhtmlpage

\subsection*{Comparative advantage and factor endowments}

A traditional statement of why comparative advantage arises is that
economies have different endowments of the factors of production -- land,
capital and labour endowments differ. A land endowment that facilitates the
harvesting of grain (Saskatchewan) or the growing of fruit (California) may
be innate to an economy. We say that wheat production is \textit{land
	intensive}, that aluminum production is \textit{power intensive}, that
research and development is \textit{skill intensive}, that auto manufacture
is \textit{capital intensive,} that apparel is \textit{labour intensive}.
Consequently, if a country is well endowed with some particular factors of
production, it is to be expected that it will specialize in producing goods
that use those inputs. A relatively abundant supply or endowment of one
factor of production tends to make the cost of using that factor relatively
cheap: It is relatively less expensive to produce clothing in China and
wheat in Canada than the other way around. This explains why Canada's
Prairies produce wheat, why Quebec produces aluminum, why Asia produces
apparel. But endowments can evolve.

How can we explain why Switzerland specializes in watches, precision
instruments, and medical equipment, while Vietnam specializes in rice and
tourism? Evidently, Switzerland made a decision to educate its population
and invest in the capital required to produce these goods. It was not
naturally endowed with these skills, in the same way that Greece is endowed
with sun or Saskatchewan is endowed with fertile flat land.

While we have demonstrated the principle of comparative advantage using a
two-good example (since we are constrained by the geometry of two
dimensions), the conclusions carry over to the case of many goods.
Furthermore, the principle has many applications. For example, if one person
in the household is more efficient at doing all household chores than
another, there are still gains to specialization provided the efficiency
differences are not all identical. This is the principle of comparative
advantage at work in a microcosm.

\newhtmlpage

\begin{ApplicationBox}{caption={The one hundred mile diet\label{app:ch15app1}}}
	In 2005 two young British Columbians embarked on what has famously become known as the `one hundred mile diet'---a challenge to eat and drink only products grown within this distance of their home. They succeeded in doing this for a whole year, wrote a book on their experience and went on to produce a TV series. They were convinced that such a project is good for humanity, partly because they wrapped up ideas on organic farming and environmentally friendly practices in the same message.

	Reflect now on the implications of this superficially attractive program: If North Americans were to espouse this diet, it would effectively result in the closing down of the midwest of the Continent. From Saskatchewan to Kansas, we are endowed with grain-producing land that is the envy of the planet. But since most of this terrain is not within 100 miles of any big cities, these deluded advocates are proposing that we close up the production of grains and cereals exactly in those locations where such production is extraordinarily efficient. Should we sacrifice grains and cereals completely in this hemisphere, or just cultivate them on a hillside close to home, even if the resulting cultivation were to be more labour and fuel intensive? Should we produce olives in greenhouses in Edmonton rather than importing them from the Mediterranean, or simply stop eating them? Should we sacrifice wine and beer in North Battleford because insufficient grapes and hops are grown locally?

	Would production in temperate climates really save more energy than the current practice of shipping vegetables and fruits from a distance---particularly when there are returns to scale associated with their distribution? The `one hundred mile diet' is based on precepts that are contrary to the norms of the gains from trade. In its extreme the philosophy proposes that food exports be halted and that the world's great natural endowments of land, water, and sun be allowed to lie fallow. Where would that leave a hungry world? 
\end{ApplicationBox}
