\section{Trade barriers: Tariffs, subsidies and quotas}\label{sec:ch15sec5}

Despite the many good arguments favoring free or relatively free trade, we
observe numerous trade barriers. These barriers come in several forms. A %
\terminology{tariff} is a tax on an imported product that is designed to
limit trade in addition to generating tax revenue. It is a barrier to trade.
There also exist \terminology{quotas}, which are quantitative restrictions
on imports; other \terminology{non-tariff barriers}, such as product content
requirements; and \terminology{subsidies}. By raising the domestic price of
imports, a tariff helps domestic producers but hurts domestic consumers.
Quotas and other non-tariff barriers have similar impacts.

\begin{DefBox}
	A \textbf{tariff} is a tax on an imported product that is designed to limit trade in addition to generating tax revenue. It is a barrier to trade.
	
	A \textbf{quota} is a quantitative limit on an imported product.
	
	A \textbf{trade subsidy} to a domestic manufacturer reduces the domestic cost and limits imports.
	
	\textbf{Non-tariff barriers}, such as product content requirements, limit the gains from trade.
\end{DefBox}

\newhtmlpage

\begin{ApplicationBox}{caption={Tariffs -- the national policy of J.A. MacDonald \label{app:ch15app2}}}
	In Canada, tariffs were the main source of government revenues, both before and after Confederation in 1867 and up to World War I. They provided `incidental protection' for domestic manufacturing. After the 1878 federal election, tariffs were an important part of the National Policy introduced by the government of Sir John A. MacDonald. The broad objective was to create a Canadian nation based on east-west trade and growth.

	This National Policy had several dimensions. Initially, to support domestic manufacturing, it increased tariff protection on foreign manufactured goods, but lowered tariffs on raw materials and intermediate goods used in local manufacturing activity. The profitability of domestic manufacturing improved. But on a broader scale, tariff protection, railway promotion, Western settlement, harbour development, and transport subsidies to support the export of Canadian products were intended to support national economic development. Although reciprocity agreements  with the United States removed duties on commodities for a time, tariff protection for manufactures was maintained until the GATT negotiations of the post-World War II era. 
\end{ApplicationBox}

\newhtmlpage

\subsection*{Tariffs}

Figure~\ref{fig:tarifftrade} describes how tariffs operate. We can think of
this as the Canadian wine market---a market that is heavily taxed in Canada.
The world price of Cabernet Sauvignon is \$10 per bottle, and this is shown
by the horizontal world supply curve at that price. It is horizontal because
our domestic market accounts for only a small part of the world demand for
wine. International producers can supply us with any amount we wish to buy
at the world price. The Canadian demand for this wine is given by the demand
curve $D$, and Canadian suppliers have a supply curve given by $S$ (Canadian
Cabernet is assumed to be of the same quality as the imported variety in
this example). At a price of \$10, Canadian consumers wish to buy $Q_{D}$
litres, and domestic producers wish to supply $Q_{S}$ litres. The gap
between domestic supply $Q_{S}$ and domestic demand $Q_{D}$ is filled by
imports. This is the \textit{free trade equilibrium}.

% Figure 15.4
\input{figures/ch15/ch15fig4}

\newhtmlpage

If the government now imposes a 20 percent tariff on imported wines, foreign
wine sells for \$12 a bottle, inclusive of the tariff. The tariff raises the
domestic `tariff-inclusive' price above the world price, and this shifts the
supply of this wine upwards. By raising wine prices in the domestic market,
the tariff protects domestic producers by raising the domestic price at
which imports become competitive. Those domestic suppliers who were
previously not quite competitive at a global price of \$10 are now
competitive. The total quantity demanded falls from $Q_D$ to $Q'_D$ 
at the new equilibrium F. Domestic producers supply the amount $Q'_S$ 
and imports fall to the amount $(Q'_D-Q'_S)$. Reduced 
imports are partly displaced by domestic
producers who can supply at prices between \$10 and \$12. Hence, imports
fall both because total consumption falls and because domestic suppliers can
displace some imports under the protective tariff.

Since the tariff is a type of tax, its impact in the market depends upon the
elasticities of supply and demand, (as illustrated in 
Chapters~\ref{chap:elasticities} and~\ref{chap:welfare}). The
more elastic is the demand curve, the more a given tariff reduces imports.
In contrast, if it is inelastic the quantity of imports declines less.

\newhtmlpage

\subsection*{Costs and benefits of a tariff}

The costs of a tariff come from the higher price to consumers, but this is
partly offset by the tariff revenue that goes to the government. This tariff
revenue is a benefit and can be redistributed to consumers or spent on goods
from which consumers derive a benefit. But there are also efficiency costs
associated with tariffs---deadweight losses, as we call them. These are the
real costs of the tariff, and they arise because the marginal cost of
production does not equal the marginal benefit to the consumer. Let us see
how these concepts apply with the help of Figure~\ref{fig:tarifftrade}.

Consumer surplus is the area under the demand curve and above the
equilibrium market price. It represents the total amount consumers would
have been willing to pay for the product but did not have to pay at the
equilibrium price. It is a measure of consumer welfare. The tariff raises
the market price and reduces this consumer surplus by the amount LFGJ. This
area measures by how much domestic consumers are worse off as a result of
the price increase caused by the tariff. But this is not the net loss for
the whole domestic economy, because the government obtains some tax revenue
and domestic producers get more revenue and profit.

Government revenue accrues from the domestic sales of imports. On imports of 
$(Q'_{D}-Q'_{S})$, tax revenue is EFHI. Then,
domestic producers obtain an additional profit of LECJ---the excess of
additional revenue over their cost per additional bottle. If we are not
concerned about who gains and who loses, then there is a net loss to the
domestic economy equal to the areas CEI and HFG.

The area HFG is the standard measure of deadweight loss. At the quantity 
$Q'_D$, the cost of an additional bottle is less than the value
placed on it by consumers; and, by not having those additional bottles
supplied, consumers forgo a potential gain. The area A tells us that when
supply by domestic higher-cost producers is increased, and supply of
lower-cost foreign producers is reduced, the corresponding resources are not
being used efficiently. The sum of the areas CEI and HFG is therefore the total
deadweight loss of the tariff.

\newhtmlpage

\subsection*{Production subsidies}

Figure~\ref{fig:subsidytrade} illustrates the effect of a subsidy to a
domestic supplier. As in Figure~\ref{fig:tarifftrade}, the amount $Q_D$ is
demanded in the free trade equilibrium and, of this, $Q_S$ is supplied
domestically. With a subsidy per unit of output sold, the government can
reduce the supply cost of the domestic supplier, thereby shifting the supply
curve downward from $S$ to $S'$. In this illustration, the total
quantity demanded remains at $Q_D$, but the domestic share increases to 
$Q'_S$.

% Figure 15.5
\input{figures/ch15/ch15fig5}

\newhtmlpage

The new equilibrium represents a misallocation of resources. When domestic
output increases from $Q_S$ to $Q'_S$, a low-cost international
producer is being replaced by a higher cost domestic supplier; the domestic
supply curve $S$ lies above the international supply curve $P$ in this range
of output.

Note that this example deals with a subsidy to domestic suppliers who are
selling in the domestic market. It is not a subsidy to domestic producers
who are selling in the international market -- an export subsidy.

This subsidy comes with a cost to the domestic economy: Taxpayers-at-large
must pay higher taxes to support this policy; and each dollar raised in tax
itself has a deadweight loss, as we examined in Chapter~\ref{chap:welfare}.

\newhtmlpage

\subsection*{Quotas}

A quota is a limit placed upon the amount of a good that can be imported.
Consider Figure~\ref{fig:quotatrade}, where again there is a domestic supply
curve coupled with a world price of $P$. Rather than imposing a tariff, the
government imposes a quota that restricts imports to a physical amount
denoted by the distance \textit{quota} on the quantity axis. The supply
curve facing domestic consumers then has several segments to it. First it
has the segment $RC$, reflecting the fact that domestic suppliers are
competitive with world suppliers up to the amount $C$. Beyond this output,
world suppliers can supply at a price of $P$, whereas domestic suppliers
cannot compete at this price. Therefore the supply curve becomes horizontal,
but only \textit{up to the amount permitted under the quota}---the quantity 
$CU$ corresponding to \textit{quota}. Beyond this amount, international
supply is not permitted and therefore additional amounts are supplied by the
(higher cost) domestic suppliers. Hence the supply curve to domestic buyers
becomes the supply curve from the domestic suppliers once again.

% Figure 15.6
\input{figures/ch15/ch15fig6}

\newhtmlpage

The resulting supply curve yields an equilibrium quantity $Q'_{D}$. 
There are several features to note about this equilibrium. First, the
quota pushes the domestic price above the world price ($P_{dom}$ is greater
than $P$) because low-cost international suppliers are partially supplanted
by higher-cost domestic suppliers. Second, if the quota is chosen
`appropriately', the same domestic market price could exist under the quota
as under the tariff in Figure~\ref{fig:tarifftrade}. Third, in contrast to
the tariff case, the government obtains no tax revenue from the quotas.
Fourth, there are inefficiencies associated with the equilibrium at 
$Q'_{D}$. These inefficiencies arise because the lower-cost
international suppliers are not permitted to supply the amount they would be
willing to supply at the quota-induced market equilibrium. In other words,
more efficient producers are being squeezed out of the market by quotas that
make space for less-efficient producers. 

\begin{ApplicationBox}{caption={Cheese quota in Canada \label{app:ch15app3}}}
	In 1978 the federal government set a cheese import quota for Canada at just over 20,000 tonnes. This quota was implemented initially to protect the interests of domestic suppliers. Despite a strong growth in population and income in the intervening decades, the import quota has remained unchanged. The result is a price for cheese that is considerably higher than it would otherwise be. The quotas are owned by individuals and companies who have the right to import cheese. The quotas are also traded among importers, at a price. Importers wishing to import cheese beyond their available quota pay a tariff of about 250 percent. So, while the consumer is the undoubted loser in this game, who gains?

	First the suppliers gain, as illustrated in Figure~\ref{fig:quotatrade}. Canadian consumers are required to pay high-cost domestic producers who displace lower-cost producers from overseas. Second, the holders of the quotas gain. With the increase in demand for cheese that comes with higher incomes, the domestic cheese price increases over time and this in turn makes an individual quota more valuable.
\end{ApplicationBox}
