\section{Institutions governing trade}\label{sec:ch15sec7}

In the nineteenth century, world trade grew rapidly, in part because the
leading trading nation at the time---the United Kingdom---pursued a vigorous
policy of free trade. In contrast, US tariffs averaged about 50 percent,
although they had fallen to around 30 percent by the early 1920s. As the
industrial economies went into the Great Depression of the late 1920s and
1930s, there was pressure to protect domestic jobs by keeping out imports.
Tariffs in the United States returned to around 50 percent, and the United
Kingdom abandoned the policy of free trade that had been pursued for nearly
a century. The combination of world recession and increasing tariffs led to
a disastrous slump in the volume of world trade, further exacerbated by
World War II.

\subsection*{The WTO and GATT}

After World War II, there was a collective determination to see world trade
restored. Bodies such as the International Monetary Fund and the World Bank
were set up, and many countries signed the General Agreement on Tariffs and
Trade (GATT), a commitment to reduce tariffs successively and dismantle
trade restrictions.

Under successive rounds of GATT, tariffs fell steadily. By 1960, United
States tariffs were only one-fifth their level at the outbreak of the War.
In the United Kingdom, the system of wartime quotas on imports had been
dismantled by the mid-1950s, after which tariffs were reduced by nearly half
in the ensuing 25 years. Europe as a whole moved toward an enlarged European
Union in which tariffs between member countries have been abolished. By the
late 1980s, Canada's tariffs had been reduced to about one-quarter of their
immediate post-World War II level.

The GATT Secretariat, now called the World Trade Organization (WTO), aims
both to dismantle existing protection that reduces efficiency and to extend
trade liberalization to more and more countries. Tariff levels throughout
the world are now as low as they have ever been, and trade liberalization
has been an engine of growth for many economies. The consequence has been a
substantial growth in world trade.

\newhtmlpage

\subsection*{NAFTA, the EU and the TPP}

In North America, recent policy has led to a free trade area that covers the
flow of trade between Canada, the United States, and Mexico. The
Canada/United States free trade agreement (FTA) of 1989 expanded in 1994 to
include Mexico in the North American Free Trade Agreement (NAFTA). The
objective in both cases was to institute free trade between these countries
in most goods and services. This meant the elimination of tariffs over a
period of years and the reduction or removal of non-tariff barriers to
trade, with a few exceptions in specific products and cultural industries. A
critical component of the Agreement was the establishment of a dispute-resolution 
mechanism, under which disputes would be resolved by a panel of
`judges' nominated from the member economies. Evidence of the success of
these agreements is reflected in the fact that Canadian exports have grown
to more than 30 percent of GDP, and trade with the United States accounts
for the lion's share of Canadian trade flows. As of 2015, most of the
Pacific Rim economies were engaged in negotiating a new trade pact -- the
Trans-Pacific Partnership.

The European Union was formed after World War II, with the prime objective
of bringing about a greater degree of \textit{political} integration in
Europe. Two world wars had laid waste to their economies and social fabric.
Closer economic ties and greater trade were seen as the means of achieving
this integration. The Union was called the ``Common
Market'' for much of its existence. The Union originally
had six member states, but as of 2015 the number is 28, with several other
candidate countries in the process of application, most notably Turkey. The
European Union (EU) has a secretariat and parliament in Bruxelles. You can
find more about the EU at \url{http://europa.eu}.