\section{Imperfect competitors}\label{sec:ch11sec1}

The preceding chapters have explored extreme forms of supply. The monopolist
is the sole supplier and possesses as much market power as possible. In
contrast the perfect competitor is small and has no market power whatsoever.
He simply accepts the price for his product that is determined in the market
by the forces of supply and demand. These are very useful paradigms to
explore, but the real world for the most part lies between these extremes.
We observe that there are a handful of dominant brewers in Canada who supply
more than three quarters of the market, and they are accompanied by numerous
micro brewers that form the fringe of the brewing business. We have a small
number of air carriers and one of them controls half of the national market.
The communications market has just three major suppliers; the Canadian
Football League has eight teams and there are just a handful of major
hardware/builders' suppliers stores nationally. At the other end of the
spectrum we have countless restaurants and fitness centres, but they do not
supply the same product to the marketplaces for `food' or `health', and so
these markets are not perfectly competitive, despite the enormous number of
participants.

In this chapter we will explore the relationship between firm behaviour and
the size of the firm relative to the whole sector.

In addition to identifying imperfect competitors by the number of firms in
their sector, or the share of total sales going to a small number of
suppliers, we can differentiate imperfect competition from perfect
competition by the characteristics of the demand curves they all face. A
perfect competitor faces a perfectly elastic demand at the existing market
price; this is the only market structure to have this characteristic. In all
other market structures suppliers effectively face a downward-sloping
demand. This means that they have some influence on the price of the good,
and also that if they change the price they charge, they can expect demand
to reflect this in a predictable manner. So we will lump all other market
structures, including the extreme of monopoly, under the title of %
\terminology{imperfect competition}, and firms in such markets face downward
sloping demands.

\begin{DefBox}
	\textbf{Imperfectly competitive firms} face a downward-sloping demand curve, and their output price reflects the quantity sold.
\end{DefBox}

\newhtmlpage

The demand curve for the firm and industry coincide for the monopolist, but
not for other imperfectly competitive firms. It is convenient to categorize
the producing sectors of the economy as either having a relatively small
number of participants, or having a large number. The former market
structures are called \terminology{oligopolistic}, and the latter are called %
\terminology{monopolistically competitive}. The word \textit{oligopoly}
comes from the Greek word \textit{oligos} meaning few, and \textit{polein}
meaning to sell.

\begin{DefBox}
	\textbf{Oligopoly} defines a market with a small number of suppliers.
	
	\textbf{Monopolistic competition} defines a market with many sellers of products that have similar characteristics. Monopolistically competitive firms can exert only a small influence on the whole market.
\end{DefBox}

The home appliance industry is an oligopoly. The prices of \textit{KitchenAid} 
appliances depend not only on its own output and sales, but also on the
prices of \textit{Whirlpool}, \textit{Maytag} and \textit{Bosch}. If a firm
has just two main producers it is called a duopoly. \textit{Canadian National} 
and \textit{Canadian Pacific} are the only two major rail freight carriers
in Canada; they form a duopoly. In contrast, the local Italian restaurant is
a monopolistic competitor. Its output is a package of distinctive menu
choices, personal service, and convenience for local customers. It can
charge a different price than the out-of-neighbourhood restaurant, but if
its prices are too high local diners may travel elsewhere for their food
experience. Many markets are defined by producers who supply similar but not
identical products. Canada's universities all provide degrees, but the
institutions differ one from another in their programs, their balance of
in-class and on-line courses, their student activities, whether they are
science based or liberal arts based, whether they have cooperative programs
or not, and so forth. While universities are not in the business of making
profit, they certainly wish to attract students, and one way of doing this
is to differentiate themselves from other institutions. The profit-oriented
world of commerce likewise seeks to increase its market share by
distinguishing its product line.

\begin{DefBox}
	\textbf{Duopoly} defines a market or sector with just two firms.
\end{DefBox}

These distinctions are not completely airtight. For example, if a sole
domestic producer is subject to international competition it cannot act in
the way we described in the previous chapter -- it has potential, or actual,
competition. \textit{Bombardier} may be Canada's sole rail car manufacturer,
but it is not a monopolist, even in Canada. It could best be described as
being part of an international oligopoly in rail-car manufacture. Likewise, it
is frequently difficult to delineate the boundary of a given market. For
example, is \textit{Canada Post} a monopoly in mail delivery, or an
oligopolist in hard-copy communication? We can never fully remove these
ambiguities.

\newhtmlpage

\subsection*{The role of cost structures}

A critical determinant of market structure is the way in which demand and
cost interact to determine the likely number of market participants in a
given sector or market. Structure also evolves over the long run: Time is
required for entry and exit.

Figure~\ref{fig:demandcostmarketstructure} shows the demand curve $D$ for
the output of an industry in the long run. Suppose, initially, that all
firms and potential entrants face the long-run average cost curve $LATC_1$.
At the price $P_1$, free entry and exit means that each firm produces $q_1$.
With the demand curve $D$, industry output is $Q_1$. The number of firms in
the industry is $N_1$ ($=Q_1/q_1$). If $q_1$, the minimum average cost
output on $LATC_1$, is small relative to $D$, then $N_1$ is large. This
outcome might be perfect competition -- $N$ virtually infinite -- or
monopolistic competition -- $N$ large with slightly differentiated products
produced by each firm.

% Figure 11.1
\input{figures/ch11/ch11fig1}

\newhtmlpage

Instead, suppose that the production structure in the industry is such that
the long-run average cost curve is $LATC_2$. Here, scale economies are vast,
relative to the market size. At the lowest point on this cost curve, output
is large relative to the demand curve $D$. If this one firm were to act like
a monopolist it would produce an output where $MR=MC$ in the long run and
set a price such that the chosen output is sold. Given the scale economies,
there may be no scope for another firm to enter this market, because such a
firm would have to produce a very high output to compete with the existing
producer. This situation is what we previously called a ``natural''
monopolist.

Finally, the cost structure might involve curves of the type $LATC_3$, which
would give rise to the possibility of several producers, rather than one or
very many. This results in oligopoly.

It is clear that one \textit{crucial determinant of market structure is
	minimum efficient scale relative to the size of the total market} as shown
by the demand curve. The larger the minimum efficient scale relative to
market size, the smaller is the number of producers in the industry.