\section{Monopolistic competition}\label{sec:ch11sec3}

Monopolistic competition presumes a large number of quite small producers or
suppliers, each of whom may have a slightly \terminology{differentiated
	product}. The competition element of this name signifies that there are many
participants, while the monopoly component signifies that each supplier
faces a downward-sloping demand. In concrete terms, your local coffee shop
that serves ``fair trade'' coffee has a product that differs slightly from
that of neighbouring shops that sell the traditional product. They coexist
in the same sector, and probably charge different prices: The fair trade
supplier likely charges a higher price, but knows nonetheless that too large
a difference between her price and the prices of her competitors will see
some of her clientele migrate to those lower-priced establishments. That is
to say, she faces a downward-sloping demand curve.

The competition part of the name also indicates that there is \textit{free
	entry and exit}. There are no barriers to entry. As a consequence, we know
at the outset that only normal profits will exist in a long-run equilibrium.
Economic profits will be competed away by entry, just as losses will erode
due to exit.

As a general rule then, each firm can influence its market share to some
extent by changing its price. Its demand curve is not horizontal because
different firms' products are only limited substitutes. A lower price level
may draw some new customers away from competitors, but convenience or taste
will prevent most patrons from deserting their local businesses. In concrete
terms: A pasta special at the local Italian restaurant that reduces the
price below the corresponding price at the competing local Thai restaurant
will indeed draw clients away from the latter, but the foods are
sufficiently different that only some customers will leave the Thai
restaurant. The differentiated menus mean that many customers will continue
to pay the higher price.

\begin{DefBox}
	A \textbf{differentiated product} is one that differs slightly from other products in the same market.
\end{DefBox}

Given that there are very many firms, the theory also envisages limits to
scale economies. Firms are small and, with many competitors, individual
firms do not compete strategically with \textit{particular} rivals. Because
the various products offered are slightly differentiated, we avoid graphics
with a \textit{market} demand, because this would imply that a uniform
product is being considered. At the same time the market is a well-defined
concept---it might be composed of all those restaurants within a reasonable
distance, for example, even though each one is slightly different from the
others. The market share of each firm depends on the price that it charges 
\textit{and} on the number of competing firms. For a given number of
suppliers, a shift in industry demand also shifts the demand facing each
firm. Likewise, the presence of more firms in the industry reduces the
demand facing each one.

\newhtmlpage

The market equilibrium is illustrated in Figure~\ref{fig:eqmonocomp}. Here 
$D_0$ is the initial demand facing a representative firm, and $MR_0$ is the
corresponding marginal revenue curve. Profit is maximized where $MC=MR$, and
the price $P_0$ is obtained from the demand curve corresponding to the
output $q_0$. Total profit is the product of output times the difference
between price and average cost, which equals $q_0\times(P_0-AC_0)$.

% Figure 11.2
\input{figures/ch11/ch11fig2}

\newhtmlpage

With free entry, such profits attract new firms. The increased number of
firms reduces the share of the market that any one firm can claim. That is,
the firm's demand curve shifts inwards when entry occurs. As long as
(economic) profits exist, this process continues. For entry to cease,
average cost must equal price. A final equilibrium is illustrated by the
combination $(P_{E},q_{E})$, where the demand has shifted inward to $D$.

At this long-run equilibrium, two conditions must hold: First, the optimal
pricing rule must be satisfied---that is $MC=MR$; second it must be the case
that only normal profits are made at the final equilibrium. Economic profits
are competed away as a result of free entry. And these two conditions must
exist at the optimal output. Graphically this implies that $ATC$ must equal
price at the output where $MC=MR$. In turn this implies that the $ATC$ is
tangent to the demand curve where $P=ATC$. While this could be proven
mathematically, it is easy to intuit why this tangency must exist: If $ATC$
merely intersected the demand curve at the output where $MC=MR$, we could
find some other output where the demand price would be above $ATC$,
suggesting that profits could be made at such an output. Clearly that could
not represent an equilibrium.

\begin{DefBox}
	The \textbf{monopolistically competitive equilibrium} in the long run requires the firm's demand curve to be tangent to the $ATC$ curve at the output where $MR=MC$.
\end{DefBox}