\section{Rationality}\label{sec:ch6sec1}

A critical behavioural assumption in economics is that agents operate in a
way that is oriented towards achieving a goal. This can be interpreted to
mean that individuals and firms maximize their personal well-being and/or
their profits. These players may have other goals in addition: Philanthropy
and the well-being of others are consistent with individual optimization.

If individuals are to achieve their goals then they must act in a manner
that will get them to their objective; broadly, they must act in a rational
manner. The theory of individual maximization that we will develop in this
chapter is based on that premise or assumption. In assuming individuals are
rational we need not assume that they have every piece of information
available to them that might be relevant for a specific decision or choice.
Nor need we assume that they have super computers in their brain when they
evaluate alternative possible strategies.

What we do need to assume, however, is that individuals act in a manner that
is consistent with obtaining a given objective. The modern theory of
behavioural economics and behavioural psychology examines decision making in
a wide range of circumstances and has uncovered many fascinating behaviours
-- some of which are developed in Application Box~\ref{app:rationalimpulse}
below.

We indicated in Chapter~\ref{chap:intro} that as social scientists, we
require a \textit{reliable model} of behaviour, that is, \textit{a way of
	describing the essentials of choice that is consistent with everyday
	observations on individual behaviour patterns}. In this chapter, our aim is
to understand more fully the behavioural forces that drive the demand side
of the economy.

Economists analyze individual decision making using two different, yet
complementary, approaches -- utility analysis and indifference analysis. We
begin by portraying individuals as maximizing their \textit{measurable
	utility} (sometimes called \textit{cardinal utility}); then progress to
indifference analysis, where a weaker assumption is made on the ability of
individuals to measure their satisfaction. In this second instance we do not
assume that individuals can measure their utility numerically, only that
they can say if one collection of goods and services yields them greater
satisfaction than another group. This ranking of choices corresponds to what
is sometimes called \textit{ordinal utility} -- because individuals can 
\textit{order} groups of goods and services in ascending order of
satisfaction. In each case individuals are perceived as rational maximizers
or optimizers: They allocate their income so as to choose the outcome that
will make them as well off as possible.

The second approach to consumer behaviour is frequently omitted in
introductory texts. It can be omitted here without interpreting the flow of
ideas, although it does yield additional insights into consumer choice and
government policy. As in preceeding chapters, we begin the analysis with 
a motivating numerical example.

\newhtmlpage

% Application Box 6.1
\begin{ApplicationBox}{caption={Rationality and impulse \label{app:rationalimpulse}}}
	A number of informative and popular books on decision making have appeared recently. Their central theme is that our decision processes should not be viewed solely as a rational computer -- operating in one single mode only, and unmoved by our emotions or history. Psychologists now know that our brains have at least two decision modes, and these are developed by economics Nobel Prize winner Daniel Kahneman in his book ``Thinking, Fast and Slow''. One part of our brain operates in a rational goal-oriented forward-looking manner (the `slow' part), another is motivated by immediate gratification (the `fast' part). Decisions that we observe in the world about us reflect these different mechanisms. 
	
	Richard Thaler, a Chicago economist and his legal colleague Cass Sunstein, have developed a role for public policy in their book entitled ``Nudge''. They too argue that individuals do not inevitably operate in their own best long-term interests, and as a consequence individuals frequently require a \emph{nudge} by government to make the long-term choice rather than the short-term choice. For example, when individuals begin a new job, they might be automatically enrolled in the company pension plan and be given the freedom to opt out, rather than not be enrolled and given the choice to opt in. Such policies are deemed to be `soft paternalism'. They are paternalistic for the obvious reason -- another organism is directing, but they are also soft in that they are not binding. In the utility maximizing framework that economists use to model behaviour, they are assuming that individuals are operating in their rational mode rather than their impulsive mode.   
\end{ApplicationBox}