\section{The perfect competition paradigm}\label{sec:ch9sec1}

A competitive market is one that encompasses a very large number of
suppliers, each producing a similar or identical product. Each supplier
produces an output that forms a small part of the total market, and the sum
of all of these individual outputs represents the production of that sector
of the economy. Florists, barber shops, corner stores and dry cleaners all
fit this description.

At the other extreme, a market that has just a single supplier is a
monopolist. For example, the National Hockey League is the sole supplier of
top-quality professional hockey games in North America; Hydro Quebec is a
monopoly electricity distributor in Quebec; Via Rail is the only supplier of
passenger rail services between Windsor, Ontario and the city of Quebec.

We use the word `paradigm' in the title to this section: It implies that we
will develop a \textit{model} of supply behaviour for a market in which
there are many small suppliers, producing essentially the same product,
competing with one-another to meet the demands of consumers.

The structures that we call perfect competition and monopoly are extremes in
the market place. Most sectors of the economy lie somewhere between these
limiting cases. For example, the market for internet services usually
contains several providers in any area -- some provide using a fibre cable,
others by satellite. The market for smart-phones is dominated by two major
players -- \textit{Apple} and \textit{Samsung}. Hence, while these markets that have a limited
number of suppliers are competitive in that they freely and fiercely compete
for the buyer's expenditure, these are not \terminology{perfectly competitive}
markets, because they do not have a very large number of suppliers.

In all of the models we develop in this chapter we will assume that the
objective of firms is to maximize profit -- the difference between revenues
and costs.

\begin{DefBox}
	A \textbf{perfectly competitive} industry is one in which many suppliers, producing an identical product, face many buyers, and no one participant can influence the market.
	
	\textbf{Profit maximization} is the goal of competitive suppliers -- they seek to maximize the difference between revenues and costs.
\end{DefBox}

\newhtmlpage

The presence of so many sellers in perfect competition means that each firm
recognizes its own small size in relation to the total market, and that its
actions have no perceptible impact on the market price for the good or
service being traded. Each firm is therefore a \textit{price taker}---in
contrast to a monopolist, who is a \textit{price setter}.

The same `smallness' characteristic was assumed when we examined the demands
of individuals earlier. Each buyer takes the price as given. He or she is
not big enough to be able to influence the price. In contrast, when
international airlines purchase or lease aircraft from \textit{Boeing} or 
\textit{Airbus}, they negotiate over the price and other conditions of
supply. The market models underlying these types of transactions are
examined in Chapter~\ref{chap:imperfectcompetition}.

Hence, when we describe a market as being perfectly competitive we do not
mean that other market types are not competitive; all market structure are
competitive in the sense that the suppliers wish to make profit, and they
produce as efficiently as possible in order to meet that goal.