\section{The firm's supply decision}\label{sec:ch9sec3}

The concept of \terminology{marginal revenue} is key to analyzing the supply
decision of an individual firm. We have used marginal analysis at several
points to date. In consumer theory, we saw how consumers balance the utility
per dollar \textit{at the margin} in allocating their budget. Marginal
revenue is the additional revenue accruing to the firm from the sale of one
more unit of output.

\begin{DefBox}
	\textbf{Marginal revenue} is the additional revenue accruing to the firm resulting from the sale of one more unit of output.
\end{DefBox}

In perfect competition, a firm's marginal revenue ($MR$) is the price of the
good. Since the price is constant for the individual supplier, each
additional unit sold at the price $P$ brings in the same additional revenue.
Therefore, $P=MR$. For example, whether a dry cleaning business launders 10
shirts or 100 shirts per day, the price charged to customers is the same.
This equality holds in no other market structure, as we shall see in the
following chapters.

\subsection*{Supply in the short run}

Recall how we defined the short run in the previous chapter: Each firm's
plant size is fixed in the short run, so too is the number of
firms in an industry. In the long run, each individual firm
can change its scale of operation, and at the same time new firms can enter
or existing firms can leave the industry.

Perfectly competitive suppliers face the choice of how much to produce at
the going market price: That is, the amount that will maximize their profit.
We abstract for the moment on how the price in the marketplace is
determined. We shall see later in this chapter that it emerges as the value
corresponding to the intersection of the supply and demand curves for the
whole market -- as described in Chapter~\ref{chap:classical}.

\newhtmlpage

The firm's $MC$ curve is critical in defining the optimal amount to supply
at any price. In Figure~\ref{fig:optoutput}, $MC$ is the firm's marginal
cost curve in the short run. At the price $P_{0}$ the optimal amount to
supply is $q_{0}$, the amount determined by the intersection of the $MC$ and
the demand. To see why, imagine that the producer chose to supply the
quantity $q_{1}$. Such an output would leave the opportunity for further
profit untapped. By producing one additional unit beyond $q_{1}$, the
supplier would get $P_{0}$ in additional revenue and incur an additional
cost that is less than $P_{0}$ in producing this unit. In fact, on every
unit between $q_{1}$ and $q_{0}$ he can make a profit, because the $MR$
exceeds the associated cost, $MC$. By the same argument, it makes no sense
to increase output beyond $q_{0}$, to $q_{2}$ for example, because the cost
of such additional units of output, $MC$, exceeds the revenue from them. 
\textit{The $MC$ therefore defines an optimal supply response}.

% Figure 9.1
\input{figures/ch9/ch9fig1}


\newhtmlpage

% Application box 9.1
\begin{ApplicationBox}{caption={The law of one price \label{app:lawoneprice}}}
	If information does not flow then prices in different parts of a market may differ and potential entrants may not know to enter a profitable market.

	Consider the fishermen off the coast of Kerala, India in the late 1990s. Their market was studied by Robert Jensen, a development economist. Prior to 1997, fishermen tended to bring their fish to their home market or port. This was cheaper than venturing to other ports, particularly if there was no certainty regarding price. This practice resulted in prices that were high in some local markets and low in others -- depending upon the daily catch. Frequently fish was thrown away in low-price markets even though it might have found a favourable price in another village's fish market.

	This all changed with the advent of cell phones. Rather than head automatically to their home port, fishermen began to phone several different markets in the hope of finding a good price for their efforts. They began to form agreements with buyers before even bringing their catch to port. Economist Jensen observed a major decline in price variation between the markets that he surveyed. In effect the `law of one price' came into being for sardines as a result of the introduction of cheap technology and the relatively free flow of information. 
\end{ApplicationBox}

\newhtmlpage

While the choice of the output $q_{0}$ is the best choice for the producer,
Figure~\ref{fig:optoutput} does not tell us anything about profit. To
determine profit we need information on costs. Accordingly, in
Figure~\ref{fig:shortrunsupply} the firm's $AVC$ and $ATC$ curves have been added to
Figure~\ref{fig:optoutput}. As explained in the previous chapter, the $ATC$
curve includes both fixed and variable cost components, and the $MC$ curve
cuts the $AVC$ and the $ATC$ at their minima.

% Figure 9.2
\input{figures/ch9/ch9fig2}

\newhtmlpage

First, note that any price below $P_{3}$, which corresponds to the minimum
of the $ATC$ curve, yields no profit, since it does not enable the producer
to cover all of his costs. This price is therefore called the 
\terminology{break-even price}. Second, any price below $P_{1}$, which
corresponds to the minimum of the $AVC$, does not even enable the producer
to cover variable costs. What about a price such as $P_{2}$, that lies
between these? The answer is that, if the supplier has already incurred some
fixed costs, he should continue to produce, provided he can cover his
variable cost. But in the long run he must cover all of his costs, fixed and
variable. Therefore, if the price falls below $P_{1}$, he should shut down,
even in the short run. This price is therefore called the %
\terminology{shut-down price}. If a price at least equal to $P_{3}$ cannot
be sustained in the long run, he should leave the industry. But at a price
such as $P_{2}$ he can cover variable costs and therefore should continue to
produce in the short run. His optimal output at $P_{2}$ is defined by the
intersection of the $P_{2}$ line with the $MC$ curve. The firm's 
\terminology{short-run supply curve} is, therefore, that portion of the $MC$
curve above the minimum of the $AVC$.

To illustrate this more concretely, consider again the example of our
snowboard producer, and imagine that he is producing in a perfectly
competitive marketplace. How should he behave in response to different
prices? Table~\ref{table:profitmaxsr} reproduces the data from Table~\ref{table:snowprodcost}.

\begin{Table}{caption={Profit maximization in the short run \label{table:profitmaxsr}},description={Output Price=\$70; Wage=\$1,000; Fixed Cost=\$3,000. The shut-down point occurs at a price of $\$34.3$, where the $AVC$ attains a minimum. Hence no production, even in the short run, takes place unless the price exceeds this value. The break-even level of output occurs at a price of $\$50$, where the $ATC$ attains a minimum.},descwidth={30em}}\footnotesize
\begin{tabu} to \linewidth {|X[1,c]X[1,c]X[1,c]X[1,c]X[1,c]X[1,c]X[1,c]X[1.2,c]|} \hline 
	\rowcolor{rowcolour}	\textbf{Labour} & \textbf{Output} & \textbf{Total} & \textbf{Average} & \textbf{Average} & \textbf{Marginal} & \textbf{Total} & \textbf{Profit} \\[-0.1em]
	\rowcolor{rowcolour}	&	&	\textbf{Revenue \$}	&	\textbf{Variable}	&	\textbf{Total Cost}	&	\textbf{Cost \$}	&	\textbf{Cost \$}	&	\\[-0.1em]
	\rowcolor{rowcolour}	&	&	&	\textbf{Cost}	&	\textbf{\$}	&	&	&	\\	\hline
	\textbf{L} & \textbf{Q} & \textbf{TR} & \textbf{AVC} & \textbf{ATC} & \textbf{MC} & \textbf{TC} & \textbf{TR-TC} \\
	\rowcolor{rowcolour}0 & 0 &  &  &  &  & 3,000  &  \\
	1 & 15 & 1,050 & 66.7 & 266.7 & 66.7 & 4,000 & $-2,950$ \\
	\rowcolor{rowcolour}2 & 40 & 2,800 & 50.0 & 125.0 & 40.0 & 5,000 & $-2,200$ \\
	3 & 70 & 4,900 & 42.9 & 85.7 & 33.3 & 6,000 & $-1,100$ \\
	\rowcolor{rowcolour}4 & 110 & 7,700 & 36.4 & 63.6 & 25.0 & 7,000 & 700 \\
	5 & 145 & 10,150 & 34.5 & 55.2 & 28.6 & 8,000 & 2,150 \\
	\rowcolor{rowcolour}	6 & 175 & 12,250 & 34.3 & 51.4 & 33.3 & 9,000 & 3,250 \\
	7 & 200 & 14,000 & 35.0 &  50.0 &  40.0 & 10,000 & 4,000 \\ 
	\rowcolor{rowcolour}	8 & 220 & 15,400 & 36.4 &  50.0 &  50.0 & 11,000 & 4,400 \\
	9 & 235 & 16,450 & 38.3 &  51.1 &  66.7 & 12,000 & 4,450 \\
	\rowcolor{rowcolour}	10 & 240 & 16,800 & 41.7 &  54.2 & 200.0 & 13,000 &	3,800 \\	\hline
\end{tabu}
\end{Table}

\newhtmlpage

\begin{DefBox}
	The \textbf{shut-down price} corresponds to the minimum value of the $AVC$ curve.
	
	The \textbf{break-even price} corresponds to the minimum of the $ATC$ curve.
	
	The firm's \textbf{short-run supply curve} is that portion of the $MC$ curve above the minimum of the $AVC$.
\end{DefBox}

Suppose that the price is \$70. How many boards should he produce? The
answer is defined by the behaviour of the $MC$ curve. For any output less
than or equal to 235, the $MC$ is less than the price. For example, at $L=9$
and $Q=235$, the $MC$ is \$66.7. At this output level, he makes a profit on
the marginal unit produced, because the $MC$ is less than the revenue he
gets (\$70) from selling it.

But, at outputs above this, he registers a loss on the marginal units
because the $MC$ exceeds the revenue. For example, at $L=10$ and $Q=240$,
the $MC$ is \$200. Clearly, 235 snowboards is the optimum. To produce more
would generate a loss on each additional unit, because the additional cost
would exceed the additional revenue. Furthermore, to produce fewer
snowboards would mean not availing of the potential for profit on additional
boards.

His profit is based on the difference between revenue per unit and cost per
unit at this output: $(P-ATC)$. Since the $ATC$ for the 235 units produced
by the nine workers is \$51.1, his profit margin is $\$70-\$51.1=\$18.9$
per board, and total profit is therefore $235 \times \$18.9 = \$4,441.5$.

Let us establish two other key outputs and prices for the producer. First,
the shut-down point is the minimum of his $AVC$ curve. Table~\ref{table:profitmaxsr}
indicates that the price must be at least \$34.3 for him
to be willing to supply \textit{any} output, since that is the value of the 
$AVC$ at its minimum. Second, the minimum of his $ATC$ is at \$50.
Accordingly, provided the price exceeds \$50, he will cover both variable
and fixed costs and make a maximum profit when he chooses an output where 
$P=MC$, above $P=\$50$. It follows that the short-run supply curve for
Black Diamond Snowboards is the segment of the $MC$ curve in 
Figure~\ref{fig:AMCcurve} above the $AVC$ curve.

Given that we have developed the individual firm's supply curve, the next
task is to develop the \underbar{industry} supply curve.

\newhtmlpage

\subsection*{Industry supply in the short run}

In Chapter~\ref{chap:classical} it was demonstrated that individual demands
can be aggregated into an industry demand by summing them horizontally. The
industry supply is obtained in exactly the same manner---by summing the
firms' supply quantities across all firms in the industry.

To illustrate, imagine we have many firms, possibly operating at different
scales of output and therefore having different short-run $MC$ curves. The 
$MC$ curves of two of these firms are illustrated in Figure~\ref{fig:industrysupply}. 
The $MC$ of A is below the $MC$ of B; therefore, B
likely has a smaller scale of plant than A. Consider first the supply
decisions in the price range $P_1$ to $P_2$. At any price between these
limits, only firm A will supply output -- firm B does not cover its $AVC$ in
this price range. Therefore, the joint contribution to industry supply of
firms A and B is given by the $MC$ curve of firm A. But once a price of $P_2$
is attained, firm B is now willing to supply. The $S_{sum}$ schedule is the
horizontal addition of their supply quantities. Adding the supplies of every
firm in the industry in this way yields the \terminology{industry supply}.

\begin{DefBox}
	\textbf{Industry supply (short run)} in perfect competition is the horizontal sum of all firms' supply curves.
\end{DefBox}

% Figure 9.3
\input{figures/ch9/ch9fig3}

\newhtmlpage



\newhtmlpage

\subsection*{Industry equilibrium}

Consider next the industry equilibrium. Since the industry supply is the sum
of the individual supplies, and the industry demand curve is the sum of
individual demands, an equilibrium price and quantity $(P_E,Q_E)$ are
defined by the intersection of these industry-level curves, as in 
Figure~\ref{fig:marketeq}. Here, each firm takes $P_E$ as given (it is so small that it
cannot influence the going price), and supplies an amount determined by the
intersection of this price with its $MC$ curve. The sum of such quantities
is therefore $Q_E$.

\begin{DefBox}
	\textbf{Short-run equilibrium} in perfect competition occurs when each firm maximizes profit by producing a quantity where $P=MC$, provided the price exceeds the minimum of the average variable cost.
\end{DefBox}


% Figure 9.4
\input{figures/ch9/ch9fig4}

