\section{Long-run industry supply}\label{sec:ch9sec5}

When aggregating the firm-level supply curves, as illustrated in 
Figure~\ref{fig:industrysupply}, we did not assume that all firms were identical. In
that example, firm A has a cost structure with a lower $AVC$ curve, since
its supply curve starts at a lower dollar value. This indicates that firm A
may have a larger plant size than firm B -- one that puts A closer to the
minimum efficient scale region of its long-run $ATC$ curve.

% Figure 9.7
\input{figures/ch9/ch9fig7}

\newhtmlpage

Can firm B survive with his current scale of operation in the long run? Our
industry dynamics indicate that it cannot. The reason is that, provided 
\textit{some} firms are making economic profits, new entrepreneurs will
enter the industry and drive the price down to the minimum of the $ATC$ curve 
\textit{of those firms who are operating with the lowest cost plant size}.
B-type firms will therefore be forced either to leave the industry or to
adjust to the least-cost plant size---corresponding to the lowest point on
its \underbar{long-run} $ATC$ curve. Remember that the same technology is
available to all firms; they each have the same long-run $ATC$ curve, and
may choose different scales of operation in the short run, as illustrated in
Figure~\ref{fig:firmdiffsize}. But in the long run they must all produce
using the minimum-cost plant size, or else they will be driven from the
market.

This behaviour enables us to define a \terminology{long-run industry supply}. 
The long run involves the entry and exit of firms, and leads to a price
corresponding to the minimum of the long-run $ATC$ curve. Therefore, if the
long-run equilibrium price corresponds to this minimum, \textit{the long-run
supply curve of the industry is defined by a particular price value---it is
horizontal at the price corresponding to the minimum of the $LATC$}. More or
less output is produced as a result of firms entering or leaving the
industry, with those present always producing at the same unit cost in a
long-run equilibrium.

\begin{DefBox}
	\textbf{Industry supply in the long run in perfect competition} is horizontal at a price corresponding to the minimum of the representative firm's long-run $ATC$ curve.
\end{DefBox}

\newhtmlpage

% Figure 9.8
\input{figures/ch9/ch9fig8}

\newhtmlpage

This industry's long-run supply curve, $S_L$, and a particular short-run
supply are illustrated in Figure~\ref{fig:lrdynamics}. Different points on 
$S_L$ are attained when demand shifts. Suppose that, from an initial
equilibrium $Q_1$, defined by the intersection of $D_1$ and $S_1$, demand
increases from $D_1$ to $D_2$ because of a growth in income. With a fixed
number of firms, the additional demand can be met only at a higher price ($P_2$),
where each existing firm produces more using their existing plant size. The
economic profits that result induce new operators to produce. This addition
to the industry's production capacity shifts the short-run supply outwards
and price declines until normal profits are once again being made. The new
long-run equilibrium is at $Q_2$, with more firms each producing at the
minimum of their long-run $ATC$ curve, $P_E$.

The same dynamic would describe the industry reaction to a decline in
demand---price would fall, some firms would exit, and the resulting
contraction in supply would force the price back up to the long-run
equilibrium level. This is illustrated by a decline in demand from $D$ to 
$D_3$.

\newhtmlpage

\subsection*{Increasing and decreasing cost industries}

While a horizontal long-run supply is the norm for perfect competition, in
some industries costs increase with the scale of industry output; in others
they decrease. This may be because all of the producers use a particular
input that itself becomes more or less costly, depending upon the amount
supplied.

% Figure 9.9
\input{figures/ch9/ch9fig9}

\newhtmlpage

\terminology{Decreasing cost} sectors are those that benefit from a decline in
the prices of their inputs as the size of their market expands. This is
frequently because the suppliers of the inputs themselves can benefit from
scale economies as a result of expansion in the market for the final good. A
case in point has been the computer market, or the tablet market: As output
in these markets has grown, the producers of videocards and random-access
memory have benefited from scale economies and thus been able to sell these
components at a lower price to the manufacturers of the final goods. An
example of an \terminology{increasing cost} market is the market for landings and
take-offs at airports. Airports are frequently limited in their ability to
expand their size and build additional runways. In such markets, as use
grows, planes about to land may have to adopt a circling holding pattern,
while those departing encounter clearance delays. Such delays increase the
time costs to passengers and the fuel and labour costs to the suppliers. 
Decreasing and increasing industry costs are reflected in the long-run 
industry supply curve by a downward-sloping segment or an upward sloping 
segment, as illustrated in Figure~\ref{fig:incdeccost}.

\begin{DefBox}
	\textbf{Increasing (decreasing) cost} industry is one where costs rise (fall) for each firm because of the scale of industry operation.
\end{DefBox}