\section{Fixed costs and sunk costs}\label{sec:ch8sec5}

The distinction between fixed and variable costs is important for producers
who encounter difficulty in making a profit. If a producer has committed
himself to setting up a plant, then he has made a decision to incur a fixed
cost. Having done this, he must now decide on a production strategy that
will maximize profit. However, the price that consumers are willing to pay
may not be sufficient to yield a profit. So, if Black Diamond Snowboards
cannot make a profit, should it shut down? The answer is that if it can
cover its variable costs, \textit{having already incurred its fixed costs},
it should stay in production, at least temporarily. By covering the variable
cost of its operation, Black Diamond is at least earning some return. A 
\terminology{sunk cost} is a fixed cost that has already been incurred and
cannot be recovered. But if the pressures of the marketplace are so great
that the total costs cannot be covered in the longer run, then this is not a
profitable business and the firm should close its doors.

Is a fixed cost always a sunk cost? No: Any production that involves capital
will incur a fixed cost -- one not varying with the amount produced. Such
capital can be financed in several ways however: It might be financed on a
very short-term lease basis, or it might have been purchased by the
entrepreneur. If it is leased on a month-to-month basis, an unprofitable
entrepreneur who can only cover variable costs (and who does not foresee
better market conditions ahead) can exit the industry quickly -- by not
renewing the lease on the capital. But an individual who has actually
purchased equipment that cannot readily be resold has essentially sunk money
into the fixed cost component of his production. This entrepreneur should
continue to produce as long as he can cover variable costs.

\begin{DefBox}
	\textbf{Sunk cost} is a fixed cost that has already been incurred and cannot be recovered, even by producing a zero output.
\end{DefBox}

\newhtmlpage

\subsection*{R \& D as a sunk cost}

Sunk costs in the modern era are frequently in the form of research and
development costs, not the cost of building a plant or purchasing machinery.
The prototypical example is the pharmaceutical industry, where it is
becoming progressively more challenging to make new drug breakthroughs --
both because the `easier' breakthroughs have already been made, and because
it is necessary to meet tighter safety conditions attaching to new drugs.
Research frequently leads to drugs that are not sufficiently effective in
meeting their target. As a consequence, the pharmaceutical sector regularly
writes off hundreds of millions of dollars of lost sunk costs -- unfruitful
research and development.

Finally, we need to keep in mind the opportunity costs of running the
business. The owner pays himself a salary, and ultimately he must recognize
that the survival of the business should not depend upon his drawing a
salary that is less than his opportunity cost. As developed in Section%
~\ref{sec:ch7sec2}, if he underpays himself in
order to avoid shutting down, he might be better off in the long run to
close the business and earn his opportunity cost elsewhere in the
marketplace.