\section{Clusters, learning by doing, scope economies}\label{sec:ch8sec8}

\subsection*{Clusters}

The phenomenon of a grouping of firms that specialize in producing related
products is called a \terminology{cluster}. For example, Ottawa has more
than its share of software development firms; Montreal has a
disproportionate share of Canada's pharmaceutical producers and electronic
game developers; Calgary has its `oil patch'; Hollywood has movies; Toronto
is Canada's financial capital. Provincial and state capitals have most of
their province's bureaucracy. Clusters give rise to externalities,
frequently in the form of ideas that flow between firms, which in turn
result in cost reductions and new products.

\begin{DefBox}
	\textbf{Cluster}: a group of firms producing similar products, or engaged in similar research.
\end{DefBox}

The most famous example of clustering is Silicon Valley, surrounding San
Francisco, in California, the original high-tech cluster. The presence of a
large group of firms with a common focus serves as a signal to workers with
the right skill set that they are in demand in such a region. Furthermore,
if these clusters are research oriented, as they frequently are, then
knowledge spillovers benefit virtually all of the contiguous firms; when
workers change employers, they bring their previously-learned skills with
them; on social occasions, friends may chat about their work and interests
and share ideas. This is a positive externality.

\newhtmlpage

\subsection*{Learning by doing}

Learning from production-related experiences frequently reduces costs: The
accumulation of knowledge that is associated with having produced a large
volume of output over a considerable time period enables managers to
implement more efficient production methods and avoid errors. We give the
term \terminology{learning by doing} to this accumulation of knowledge.

Examples abound, but the best known may be the continual improvement in the
capacity of computer chips, whose efficiency has doubled about every
eighteen months for several decades -- a phenomenon known as Moore's Law. As 
\textit{Intel Corporation} continues to produce chips it learns how to
produce each succeeding generation of chips at lower cost. Past experience
is key. Economies of scale and learning by doing therefore may not be
independent: Large firms usually require time to grow or to attain a
dominant role in their market, and this time and experience enables them to
produce at lower cost. This lower cost in turn can solidify their market
position further.

\begin{DefBox}
	\textbf{Learning by doing} can reduce costs. A longer history of production enables firms to accumulate knowledge and thereby implement more efficient production processes.
\end{DefBox}

\subsection*{Economies of scope}

\terminology{Economies of scope} refers to the possibility that it may be
less expensive to produce and sell a line of related goods than to produce
just one product from such a line. Scope economies, therefore, define the
returns or cost reductions associated with broadening a firm's product range.

Corporations like Proctor and Gamble do not produce a single product in
their health line; rather, they produce first aid, dental care, and baby
care products. Cable companies offer their customers TV, high-speed
Internet, and telephone services either individually or packaged.

\begin{DefBox}
	\textbf{Economies of scope} occur if the unit cost of producing particular products is less when combined with the production of other products than when produced alone.
\end{DefBox}

\section*{Conclusion}

Efficient production is critical to the survival of firms. Firms that do not
adopt the most efficient production methods are likely to be left behind by
their competitors. Efficiency translates into cost considerations, and the
structure of costs in turn has a major impact on market type. Some sectors
of the economy have very many firms (the restaurant business
or the dry-cleaning business), whereas other sectors have few (internet
providers or airlines). We will see in the following chapters how market
structures depend critically upon the concept of scale economies that we
have developed here.