\section{What's it all about?}\label{sec:ch1sec1}

\subsection*{The big issues}

Economics is the study of human behaviour. Since it uses scientific methods
it is called a social science. We study human behaviour to better understand
and improve our world. During his acceptance speech, a recent Nobel Laureate
in Economics suggested:

\begin{quote}
	\textit{Economics, at its best, is a set of ideas and methods for the
		improvement of society. It is not, as so often seems the case today, a set
		of ideological rules for asserting why we cannot face the challenges of
		stagnation, job loss and widening inequality.}
	
	Christopher Sims, Nobel Laureate in Economics 2011
\end{quote}

This is an elegant definition of economics and serves as a timely caution
about the perils of ideology. Economics evolves continuously as current
observations and experience provide new evidence about economic behaviour
and relationships. Inference and policy recommendations based on earlier
theories, observations and institutional structures require constant
analysis and updating if they are to furnish valuable responses to changing
conditions and problems.

Much of today's developed world faces severe challenges as a result of the
financial crisis that began in 2008. Unemployment rates among young people
are at historically high levels in several economies, government balance
sheets are in disarray, and inequality is on the rise. In addition to the
challenges posed by this severe economic cycle, the world simultaneously
faces structural upheaval: Overpopulation, climate change, political
instability and globalization challenge us to understand and modify our
behaviour.

These challenges do not imply that our world is deteriorating. Literacy
rates have been rising dramatically in the developing world for decades;
child mortality has plummeted; family size is a fraction of what it was 50
years ago; prosperity is on the rise in much of Asia; life expectancy is
increasing universally and deaths through wars are in a state of long-term
decline.

\newhtmlpage

These developments, good and bad, have a universal character and affect
billions of individuals. They involve an understanding of economies as large
organisms with interactive components. The study of economies as large
interactive systems is called \terminology{macroeconomics}. Technically,
macroeconomics approaches the economy as a complete system with interactions
and feedback effects among sectors that determine national economic
performance. These interactions and feedbacks within the system mean that
the operation of the macro-economy is more complex than the operation of the
sum of its parts. 

Application Box~\ref{app:paradoxofthrift} gives an example.

\begin{DefBox}
	\textbf{Macroeconomics}: the study of the economy as a system in which interactions and feedbacks among sectors determine national output, employment and prices.
\end{DefBox}

\newhtmlpage

\subsection*{Individual behaviours}

Economic actions, at the level of the person or organization, form the
subject matter of microeconomics. Formally, \terminology{microeconomics} is
the study of individual behaviour in the context of scarcity. Not all
individual behaviours are motivated by self-interest; many are motivated by
a concern for the well being of society-at-large. Philanthropic societies
are goal-oriented and seek to attain their objectives in an efficient
manner. 

\begin{DefBox}
	\textbf{Microeconomics}: the study of individual behaviour in the context of scarcity.
\end{DefBox}

Individual economic decisions need not be world-changing events, or
motivated by a search for profit. Microeconomics is also about how we choose
to spend our time and money. There are quite a few options to choose from:
Sleep, work, study, food, shelter, transportation, entertainment, recreation
and so forth. Because both time and income are limited we cannot do all
things all the time. Many choices are routine or are driven by necessity.
You have to eat and you need a place to live. If you have a job you have
committed some of your time to work, or if you are a student some of your
time is committed to lectures and study. There is more flexibility in other
choices. Critically, microeconomics seeks to understand and explain how we
make choices and how those choices affect our behaviour in the workplace,
the marketplace, and society more generally.

A critical element in making choices is that there exists a \textit{scarcity}
of time, or income or productive resources. Decisions are invariably subject
to limits or constraints, and it is these constraints that make decisions
both challenging and scientific.

\newhtmlpage

Microeconomics also concerns business choices. How does a business use its
funds and management skill to produce goods and services? The individual
business operator or firm has to decide what to produce, how to produce it,
how to sell it and in many cases, how to price it. To make and sell pizza,
for example, the pizza parlour needs, in addition to a source of pizza
ingredients, a store location (land), a pizza oven (capital), a cook and a
sales person (labour). Payments for the use of these inputs generate income
to those supplying them. If revenue from the sale of pizzas is greater than
the costs of production, the business earns a profit for the owner. A
business fails if it cannot cover its costs.

In these micro-level behaviours the decision makers have a common goal: To
do as well as they can, \textit{given the constraints imposed by the 
operating environment}. The individual wants to mix work and leisure in a
way that makes her as happy or contented as possible. The entrepreneur aims
at making a profit. These actors, or agents as we sometimes call them, are 
\textit{maximizing}. Such maximizing behaviour is a central theme in this
book and in economics at large.

\newhtmlpage

\begin{ApplicationBox}{caption={Macroeconomic interactions -- the paradox of thrift \label{app:paradoxofthrift}}}
	Finance Minister Jim Flaherty and Bank of Canada Governor Mark Carney in 2011 urged Canadian households to increase their savings in order to reduce their record high debt-to-income ratio. On an individual level this makes obvious sense. If you could save more and spend less you could pay down the balances on credit cards, your line of credit, mortgage and other debts.
	
	But one household's spending is another household's income. For the economy as a system, an increase in households' saving from say 5 percent of income to 10 percent reduces spending accordingly. But lower spending by all households will reduce the purchases of goods and services produced in the economy, and therefore has the potential to reduce national incomes. Furthermore, with lower income the troublesome debt-to-income ratio will not fall, as originally intended. Hence, while higher saving may work for one household in isolation, higher saving by all households may not. The interactions and feedbacks in the economic system create a `\textbf{paradox of thrift}'.
	
	The paradox can also create problems for government finances and debt. Following the recession that began in 2008/09, many European economies with high debt loads cut spending and increased taxes to in order to balance their fiscal accounts. But this fiscal austerity reduced the national incomes on which government tax revenues are based, making deficit and debt problems even more problematic. Feedback effects, within and across economies, meant that European Union members could not all cut deficits and debt simultaneously.
\end{ApplicationBox}

\newhtmlpage

\subsection*{Markets and government}

Markets play a key role in coordinating the choices of individuals with the
decisions of business. In modern market economies goods and services are
supplied by both business and government. Hence we call them 
\terminology{mixed economies}. Some products or services are available through the marketplace to
those who wish to buy them and have the necessary income---as in cases like
coffee and wireless services. Other services are provided to all people
through government programs like law enforcement and health care.

\begin{DefBox}
	\textbf{Mixed economy}: goods and services are supplied both by private suppliers and government.
\end{DefBox}

Markets offer the choice of a wide range of goods and services at various
prices. Individuals can use their incomes to decide the pattern of
expenditures and the bundle of goods and services they prefer. Businesses
sell goods and services in the expectation that the market price will cover
costs and yield a profit.

The market also allows for specialization and separation between production
and use. Rather than each individual growing her own food, for example, she
can sell her time or labour to employers in return for income. That income
can then support her desired purchases. If businesses can produce food more
cheaply than individuals the individual obviously gains from using the
market -- by both having the food to consume, and additional income with
which to buy other goods and services. Economics seeks to explain how
markets and specialization might yield such gains for individuals and
society.

We will represent individuals and firms by envisaging that they have
explicit objectives -- to maximize their happiness or profit. However, this
does not imply that individuals and firms are concerned only with such
objectives. On the contrary, much of microeconomics and macroeconomics
focuses upon the role of government: How it manages the economy through
fiscal and monetary policy, how it redistributes through the tax-transfer
system, how it supplies information to buyers and sets safety standards for
products.

\newhtmlpage

Since governments perform all of these society-enhancing functions, in
large measure governments reflect the social ethos of voters. So, while
these voters may be maximizing at the individual level in their everyday
lives, and our models of human behaviour in microeconomics certainly
emphasize this optimization, economics does not see individuals and
corporations as being devoid of civic virtue or compassion, nor does it
assume that only market-based activity is important. Governments play a
central role in modern economies, to the point where they account for more
than one third of all economic activity in the modern mixed economy.

Governments supply goods and services in many spheres, for example, health
and education. The provision of public education is motivated both by a
concern for equality and a realization that an educated labour force
increases the productivity of an economy. Likewise, the provision of law and
order, through our legal system broadly defined, represents more than a
commitment to a just society at the individual level; without a legal system
that enforces contracts and respects property rights, the private sector of
the economy would diminish dramatically as a result of corruption,
uncertainty and insecurity. It is the lack of such a secure environment in
many of the world's economies that inhibits their growth and prosperity.

Let us consider now the methods of economics, methods that are common to
science-based disciplines.