\section{Aggregate output, growth and business cycles}\label{sec:ch1sec6}

The $PPF$ can be used to illustrate several aspects of macroeconomics: In
particular, the level of an economy's output, the growth of national and per
capita output over time, and short-run business-cycle fluctuations in
national output and employment.

\subsection*{Aggregate output}

An economy's capacity to produce goods and services depends on its endowment
of resources and the productivity of those resources. The two-person,
two-product examples in the previous section reflect this.

The \terminology{productivity of labour}, defined as output per worker or
per hour, depends on:

\begin{itemize}
	\item Skill, knowledge and experience of the labour force;
	
	\item \terminology{Capital stock}: Buildings, machinery, equipment, and
	software the labour force has to work with; and
	
	\item Current state of technology.
\end{itemize}

\begin{DefBox}
	The \textbf{productivity of labour} is the output of goods and services per worker.
	
	An economy's \textbf{capital stock} is the buildings, machinery, equipment and software used in producing goods and services.
\end{DefBox}

The economy's output, which we define by $Y$, can be defined as the output
per worker times the number of workers; hence, we can write:

\[
Y=(\text{number of workers employed})\times(\text{output per worker}).
\]

\newhtmlpage

When the employment of labour corresponds to `full employment' in the sense
that everyone willing to work at current wage rates and normal hours of work
is working, the economy's actual output is also its capacity output $Y_c$.
We also term this capacity output as \terminology{full employment output}:

\begin{DefBox}
	\textbf{Full employment output} $Y_c=(\text{number of workers at full employment})\times(\text{output per worker})$.
\end{DefBox}

\newhtmlpage

Suppose the economy is operating with full employment of resources producing
outputs of two types: Goods and services. In Figure~\ref{fig:growthandppf}, 
$PPF_{0}$ shows the different combinations of goods and services that the
economy can produce in a particular year using all its labour, capital and
the best technology available at the time.

% Figure 1.5
\input{figures/ch1/ch1fig5}

\newhtmlpage

An aggregate economy produces a large variety of outputs in two broad
categories. Goods are the products of the agriculture, forestry, mining,
manufacturing and construction industries. Services are provided by the
wholesale and retail trade, transportation, hospitality, finance, health
care, education, legal and other service sectors. As in the two-product
examples used earlier, the shape of the $PPF$ illustrates the opportunity
cost of increasing the output of either product type. We are not concerned
with who supplies the products for the moment: It may be the private sector
or the government. 

Point $X_0$ on $PPF_0$ shows one possible structure of capacity output. This
combination may reflect the pattern of demand and hence expenditures in this
economy. Output structures differ among economies with different income
levels. High-income economies spend more on services than goods and produce
higher ratios of services to goods. Middle income countries produce lower
ratios of services to goods, and low income countries much lower ratios of
services to goods. Different countries also have different $PPF$s and
different output structures, depending on their labour forces, labour
productivity and expenditure patterns.

\subsection*{Economic growth}

Three things contribute to growth in the economy. The labour supply grows as
the population expands; the stock of capital grows as spending by business
(and government) on buildings, machinery, information technology and so
forth increases; and labour-force productivity grows as a result of experience, the
development of scientific knowledge combined with product and process
innovations, and advances in the technology of production. Combined, these
developments expand capacity output over time. In Figure~\ref{fig:growthandppf} 
economic growth shifts the $PPF$ out from $PPF_{0}$ to $PPF_{1}$.

This basic description covers the key sources of growth in total output.
Economies differ in their rates of overall economic growth as a result of
different rates of growth in labour force, in capital stock, and
improvements in technology. But improvements in standards of living require
more than growth in total output. Increases in output \textit{per worker}
and \textit{per person} are necessary. Sustained increases in living
standards require sustained growth in labour productivity, which in turn is
based on advances in the technology along with the amount of capital each
worker has to work with.

\newhtmlpage

\subsection*{Recessions and booms}

A prime objective of economic policy is to ensure that the economy operates
on or near the $PPF$ -- it should use its resources to capacity and have
minimal unemployment. However, economic conditions are seldom tranquil for
long periods of time. Unpredictable changes in business expectations of
future profits, in consumer confidence, in financial markets, in commodity
and energy prices, in the demand conditions in major trading partners, in
government policy and many other events disrupt patterns of expenditure and
output. Some of these changes disturb the level of total expenditure and
thus the demand for total output. Others disturb the conditions of
production and thus the economy's production capacity. Whatever the exact
cause, the economy may be pushed off its current $PPF$. If expenditures on
goods and services decline, the economy may experience a %
\terminology{recession}. Output would fall short of capacity output and
unemployment would rise. Alternatively, times of rapidly growing expenditure
and output may result in an economic \terminology{boom}: Output and
employment expand beyond capacity levels.

\begin{DefBox}
	An \textbf{economic recession} occurs when output falls below the economy's capacity output.
	
	A \textbf{boom} is a period of high growth that raises output above normal capacity output. 
\end{DefBox}

Recent history provides examples. Following the financial crisis of 2008-09
that hit the US and many other developed economies, many economies were
pushed into recessions. Expenditure on new residential construction
collapsed for lack of income and secure financing, as did business
investment, consumption spending and exports. Lower expenditures reduced
producers' revenues, forcing cuts in output and employment and reducing
household incomes. Lower incomes led to further cutbacks in spending. In
Canada in 2009 aggregate output declined by 2.9 percent, employment declined
by 1.6 percent and the unemployment rate rose from 6.1 percent in 2008 to
8.3 percent by 2010. The world's economies have been slow to recover, and
even by 2015 the output in several developed economies was no higher than it
was in 2008. Canada's recession was not nearly as severe as the recessions
in economies such as Spain and Italy; but output between 2009 and 2015 has
been far below the potential of the Canadian economy. The unemployment rate
in 2015 still stood at 7.0\%.

\newhtmlpage

% Figure 1.6
\input{figures/ch1/ch1fig6}

An economy in a recession is operating inside its $PPF$. The fall in output
from $X$ to $Z$ in Figure~\ref{fig:growthrecessions} illustrates the effect
of a recession. Expenditures on goods and services have declined. Output is
less than capacity output, unemployment is up and some plant capacity is
idle. Labour income and business profits are lower. More people would like
to work and business would like to produce and sell more output, but it
takes time for interdependent product, labour and financial markets in the
economy to adjust and increase employment and output. Monetary and fiscal
policy may be productive in specific circumstances, to stimulate demand,
increase output and employment and move the economy back to capacity output
and full employment. The development and implementation of such policies
form the core of macroeconomics.

\newhtmlpage

Alternatively, an unexpected increase in demand for exports would increase
output and employment. Higher employment and output would increase incomes
and expenditure, and in the process spread the effects of higher output
sales to other sectors of the economy. The economy would move outside its $%
PPF$, for example to $W$ in Figure~\ref{fig:growthrecessions}, by using its
resources more intensively than normal. Unemployment would fall and overtime
work would increase. Extra production shifts would run plant and equipment
for longer hours and work days than were planned when it was designed and
installed. Output at this level may not be sustainable, because shortages of
labour and materials along with excessive rates of equipment wear and tear
would push costs and prices up. Again, we will examine how the economy
reacts to such a state in our macroeconomic analysis.

Output and employment in the Canadian economy over the past twenty years
fluctuated about growth trend in the way Figure~\ref{fig:growthrecessions}
illustrates. For several years prior to 2008 the Canadian economy operated
slightly above its capacity; but once the recession arrived monetary and
fiscal policy were used to fight it -- to bring the economy back from a
point such as $Z$ towards a point such as $X$ on the $PPF$.

\subsection*{Macroeconomic models and policy}

The $PPF$ diagrams illustrate the main dimensions of macroeconomics:
Capacity output, growth in capacity output and business cycle fluctuations
in actual output relative to capacity. But these diagrams do not offer
explanations and analysis of macroeconomic activity. We need a macroeconomic 
\textit{model} to understand and evaluate the causes and consequences of
business cycle fluctuations. As we shall see, these models are based on
explanations of expenditure decisions by households and business, financial
market conditions, production costs and producer pricing decisions at
different levels of output. Models also capture the objectives of fiscal and
monetary policies and provide a framework for policy evaluation. A full
macroeconomic model integrates different sector behaviours and the feedbacks
across sectors that can moderate or amplify the effects of changes in one
sector on national output and employment.

\section*{Conclusion}

We have covered a lot of ground in this introductory chapter. It is intended
to open up the vista of economics to the new student in the discipline.
Economics is powerful and challenging, and the ideas we have developed here
will serve as conceptual foundations for our exploration of the subject. 