\section{Equity and efficiency}\label{sec:ch5sec1}

In modern mixed economies, markets and governments together determine the
output produced and also who benefits from that output. In this chapter we
explore a very broad question that forms the core of welfare economics: Are
markets a good way to allocate scarce resources in view of the fact that
they not only give rise to inequality and poverty, but also fail to capture
the impacts of productive activity on non-market participants? Mining
impacts the environment, traffic results in road fatalities, alcohol and
tobacco cause premature deaths, and prescription pills are abused. These
products all generate secondary impacts beyond their stated objective. We
frequently call these external effects. The analysis of markets in this
larger sense involves not just economic efficiency; public policy
additionally has a normative content because policies can impact the various
participants in different ways and to different degrees. %
\terminology{Welfare economics}, therefore, deals with both normative and
positive issues.

\begin{DefBox}
\textbf{Welfare economics} assesses how well the economy allocates its scarce resources in accordance with the goals of efficiency and equity.
\end{DefBox}

\newhtmlpage

Political parties on the left and right disagree on how well a market
economy works. Canada's New Democratic Party emphasizes the market's
failings and the need for government intervention, while the Progressive
Conservative Party believes, broadly, that the market fosters choice,
incentives, and efficiency. What lies behind this disagreement? The two
principal factors are \terminology{efficiency} and \terminology{equity}.
Efficiency addresses the question of how well the economy's resources are
used and allocated. In contrast, equity deals with how society's goods and
rewards are, and should be, distributed among its different members, and how
the associated costs should be apportioned.

\begin{DefBox}
\textbf{Equity} deals with how society's goods and rewards are, and should be, distributed among its different members, and how the associated costs should be apportioned.

\textbf{Efficiency} addresses the question of how well the economy's resources are used and allocated.
\end{DefBox}

Equity is also concerned with how different generations share an economy's
productive capabilities: More investment today makes for a more productive
economy tomorrow, but more greenhouse gases today will reduce environmental
quality tomorrow. These are inter-generational questions.

Climate change caused by global warming forms one of the biggest challenges
for humankind at the present time. As we shall see in this chapter,
economics has much to say about appropriate policies to combat warming.
Whether pollution-abatement policies should be implemented today or down the
road involves considerations of equity between generations. Our first task
is to develop an analytical tool which will prove vital in assessing and
computing welfare benefits and costs -- economic surplus.