\section{Market failures -- externalities}\label{sec:ch5sec5}

The consumer and producer surplus concepts we have developed are extremely
powerful tools of analysis, but the world is not always quite as
straightforward as simple models indicate. For example, many suppliers
generate pollutants that adversely affect the health of the population, or
damage the environment, or both. The term \terminology{externality} is used
to denote such impacts. Externalities impact individuals who are not
participants in the market in question, and the effects of the externalities
may not be captured in the market price. For example, electricity-generating
plants that use coal reduce air quality, which, in turn, adversely impacts
individuals who suffer from asthma or other lung ailments. While this is an
example of a negative externality, externalities can also be positive.

\begin{DefBox}
An \textbf{externality} is a benefit or cost falling on people other than those involved in the activity's market. It can create a difference between private costs or values and social costs or values.
\end{DefBox}

\newhtmlpage

We will now show why markets characterized by externalities are not
efficient, and also show how these externalities might be corrected or
reduced. The essence of an externality is that it creates a divergence
between private costs/benefits and social costs/benefits. If a steel
producer pollutes the air, and the steel buyer pays only the costs incurred
by the producer, then the buyer is not paying the full ``social'' cost of
the product. The problem is illustrated in Figure~\ref{fig:negextineff}.

% Figure 5.5	fig:negextineff
\input{figures/ch5/ch5fig5}

\newhtmlpage

\subsection*{Negative externalities}

In Figure~\ref{fig:negextineff}, the supply curve $S$ represents the cost to
the supplier, whereas $S_f$ (the \textit{full} cost) reflects, in
addition, the cost of bad air to the population. Of course, we are assuming
that this external cost is ascertainable, in order to be able to
characterize $S_f$ accurately. Note also that this illustration assumes
that, as power output increases, the external cost \textit{per unit} rises,
because the difference between the two supply curves increases with output.
This implies that low levels of pollution do less damage per unit: Perhaps
the population has a natural tolerance for low levels, but higher levels
cannot be tolerated easily and so the cost per unit is greater.

Despite the externality, \textit{an efficient level of production can still
be defined}. It is given by $Q^*$, not $Q_0$. To see why, consider the
impact of reducing output by one unit from $Q_0$. At $Q_0$ the willingness
of buyers to pay for the marginal unit supplied is $E_0$. The (private)
supply cost is also $E_0$. But from a societal standpoint there is a
pollution/health cost of A$E_0$ associated with that unit of production. The
full cost, as represented by $S_f$, exceeds the buyer's valuation.
Accordingly, if the last unit of output produced is cut, society gains by
the amount A$E_0$, because the cut in output reduces the excess of true cost
over value.

Applying this logic to each unit of output between $Q_0$ and $Q^*$, it is
evident that society can increase its well-being by the dollar amount equal
to the area $E^*$A$E_0$, as a result of reducing production.

Next, consider the consequences of reducing output further from $Q^*$.
Note that some pollution is being created here, and environmentalists
frequently advocate that pollution should be reduced to zero. However, an
efficient outcome may not involve a zero level of pollution! If the
production of power were reduced below $Q^*$, the loss in value to
buyers, as a result of not being able to purchase the good, would exceed the
full cost of its production.

If the government decreed that, instead of producing $Q^*$, no pollution
would be tolerated, then society would forgo the possibility of earning the
total real surplus equal to the area U$E^*$K. Economists do not advocate
such a zero-pollution policy; rather, we advocate a policy that permits a
``tolerable'' pollution level -- one that still results in net benefits to
society. In this particular example, the total cost of the tolerated
pollution equals the area between the private and full supply functions, K$%
E^*$VR.

\newhtmlpage

As a matter of policy, how is this market influenced to produce the amount $%
Q^*$ rather than $Q_0$? One option would be for the government to intervene
directly with production quotas for each firm. An alternative would be to
impose a \terminology{corrective tax} on the good whose production causes
the externality: With an appropriate increase in the price, consumers will
demand a reduced quantity. In Figure~\ref{fig:negextineff} a tax equal to
the dollar value V$E^*$ would shift the supply curve upward by that amount
and result in the quantity $Q^*$ being traded.

\begin{DefBox}
A \textbf{corrective tax} seeks to direct the market towards a more efficient output.
\end{DefBox}

We are now venturing into the field of environmental policy, and this is
explored in the following section. The key conclusion of the foregoing
analysis is that an efficient working of the market continues to have
meaning in the presence of externalities. An efficient output level still
maximizes economic surplus where surplus is correctly defined.

\newhtmlpage

\subsection*{Positive externalities}

Externalities of the \textit{positive} kind enable individuals or producers
to get a type of `free ride' on the efforts of others. Real world examples
abound: When a large segment of the population is inoculated against
disease, the remaining individuals benefit on account of the reduced
probability of transmission.

A less well recognized example is the benefit derived by many producers
world-wide from research and development (R\&D) undertaken in advanced
economies and in universities and research institutes. The result is that
society at large, including the corporate sector, gain from this enhanced
understanding of science, the environment, or social behaviours.

The free market may not cope any better with these positive externalities
than it does with negative externalities, and government intervention may be
beneficial. Furthermore, firms that invest heavily in research and
development would not undertake such investment if competitors could have a
complete free ride and appropriate the fruits. This is why \textit{patent
laws} exist, as we shall see later in discussing Canada's competition
policy. These laws prevent competitors from copying the product development
of firms that invest in R\&D. If such protection were not in place, firms
would not allocate sufficient resources to R\&D, which is a real engine of
economic growth. In essence, the economy's research-directed resources would
not be appropriately rewarded, and thus too little research would take place.

\newhtmlpage

While patent protection is one form of corrective action, subsidies are
another. We illustrated above that an appropriately formulated tax on a good
that creates negative externalities can reduce demand for that good, and
thereby reduce pollution. A subsidy can be thought of as a negative tax, and
can stimulate the supply of goods and services that have positive
externalities. Consider the example in Figure~\ref{fig:positiveext}.

Individuals have a demand for flu shots given by $D$. This reflects their
private valuation -- their personal willingness to pay. But the social value
of flu shots is greater. When a given number of individuals are inoculated,
the probability that others will be infected falls. Additionally, with
higher rates of inoculation, the health system will incur fewer costs in
treating the infected. Therefore, the value to society of any quantity of
flu shots is greater than the sum of the values that individuals place on
them.

% Figure 5.6	fig:positiveext
\input{figures/ch5/ch5fig6}

\newhtmlpage



Let $D_f$ reflects the full social value of any quantity of flu shots. If $%
S$ is the supply curve, the socially optimal, efficient, market outcome is
$Q^*$. How can we influence the market to move from $Q_0$ to $Q^*$?
One solution is a subsidy that would reduce the price from $P_0$ to $P^*$.
Rather than shifting the supply curve upwards, as a tax does,
the subsidy would shift the supply \textit{downward}, sufficiently to
intersect $D$ at the output $Q^*$. In some real world examples, the
value of the positive externality is so great that the government may decide
to drive the price to zero, and thereby provide the inoculation at a zero
price. For example, children typically get their MMR shots (measles, mumps,
and rubella) free of charge. Graphically, this case would have the
intersection between $S$ and $D_f$ occuring to the right of the $D$
intercept on the horizontal axis.
