\section{Environmental policy and climate change}\label{sec:ch5sec7}

\subsection*{Greenhouse gases}

The greatest externality challenge in the modern world is to control our
emissions of greenhouse gases.The emission of \terminology{greenhouse gases}
(GHGs) is associated with a wide variety of economic activities such as
coal-based power generation, oil-burning motors, wood-burning stoves, etc.
The most common GHG is carbon dioxide. The gases, upon emission, circulate
in the earth's atmosphere and, if their build-up is excessive, prevent
sufficient radiant heat from escaping. The result is a slow warming of the
earth's surface and air temperatures. It is envisaged that such temperature
increases will, in the long term, increase water temperatures, possibly
cause glacial melting, with the result that water levels worldwide will
rise. In addition to the higher water levels, which the Intergovernmental
Panel on Climate Change (IPCC) estimates will
be between one foot and one metre by the end of the 21\textsuperscript{st}
century, oceans may become more acidic, weather patterns may change and
weather events may become more variable and severe. The changes will be
latitude-specific and vary by economy and continent, and ultimately will
impact the agricultural production abilities of certain economies.

\begin{DefBox}
\textbf{Greenhouse gases} that accumulate excessively in the earth's atmosphere prevent heat from escaping and lead to \textbf{global warming}.
\end{DefBox}

While most scientific findings and predictions are subject to a degree of
uncertainty, there is little disagreement in the scientific community on the
very long-term impact of increasing GHGs in the atmosphere. There is some
skepticism as to whether the generally higher temperatures experienced in
recent decades are completely attributable to anthropogenic activity since
the industrial revolution, or whether they also reflect a natural cycle in
the earth's temperature. But scientists agree that a continuance of the
recent rate of GHG emissions will ultimately lead to serious climatic
problems. And since GHG emissions are strongly correlated with economic
growth, the very high rate of economic growth in many large-population
economies such as China and India mean that GHGs could accumulate at a
faster rate than considered likely in the 1990s.

This is an area where economic, atmospheric and environmental \textit{models}
are used to make predictions. Scientists attempt to infer something about
the relationship between temperature and climate on the one hand and carbon
dioxide concentrations in the atmosphere on the other, using historical
data. Data values are inferred by examining ice cores and tree rings from
eons past. Accordingly, there is a degree of uncertainty regarding the
precise impact of GHG concentrations on water levels, temperatures, and
extreme weather events. The consensus is that, in the presence of such
uncertainty, a wise strategy would involve controls on the further buildup
of gases.

\newhtmlpage

\subsection*{GHGs as a common property}

A critical characteristic of GHGs is that they are what we call in economics
a `common property': Every citizen in the world `owns' them, every citizen
has equal access to them, and it matters little where these GHGs originate.
Consequently, if economy A reduces its GHG emissions, economy B may simply
increase its emissions rather than incur the cost of reducing them. Hence,
economy A's behaviour goes unrewarded. This is the crux of international
agreements -- or disagreements. Since GHGs are a common property, in order
for A to have the incentive to reduce emissions, it needs to know that B
will act correspondingly.

\subsection*{The Kyoto Protocol}

The world's first major response to climate concerns came in the form of the
United Nations--sponsored Earth Summit in Rio de Janeiro in 1992. This was
followed by the signing of the Kyoto Protocol in 1997, in which a group of
countries committed themselves to reducing their GHG emissions relative to
their 1990 emissions levels by the year 2012. Canada's Parliament
subsequently ratified the Kyoto Protocol, and thereby agreed to meet
Canada's target of a 6 percent reduction in GHGs relative to the amount
emitted in 1990.

On a per-capita basis, Canada is one of the world's largest contributors to
global warming, even though Canada's percentage of the total is just 2
percent. Many of the world's major economies refrained from signing the
Protocol---most notably China, the United States, and India. Canada's
emissions in 1990 amounted to approximately 600 giga tonnes (Gt) of carbon
dioxide; but by the time we ratified the treaty in 2002, emissions were
about 25\% above that level. Hence the signing was somewhat meaningless, in
that Canada had virtually a zero possibility of attaining its target.

The target date of 2012 has come and gone; and the leaders of the world
economy, at their meeting in Copenhagen failed to come up with a new
agreement that would have greater force. In 2012 the Rio+20 summit was held
-- in Rio once again, with the objective of devising a means of reducing GHG
emissions. Several other meetings have taken place since then, but the world
still has no binding mechanism in place that will reduce GHGs to a
predictable level by a specific date.

The central challenge in this area is that developed economies are those
primarily responsible for the buildup of GHGs in the post industrial
revolution era. Developing economies, however, do not accept that the
developed economies should be free to continue to emit GHGs at current
levels, while the developing economies should be required to limit theirs at
a much lower level. To compound difficulties, there exists skepticism in
some economies regarding the urgency to implement limits on the growth in
emissions.

\newhtmlpage

\subsection*{Canada's GHG emissions}

An excellent summary source of data on Canada's emissions and performance
during the period 1990-2013 is available on Environment Canada's web site.
See:

\url{https://www.ec.gc.ca/ges-ghg/default.asp?lang=En\&n=5B59470C-1}

Canada, like many economies, has become more efficient in its use of energy
(the main source of GHGs) in recent decades---its \textit{use of energy per
unit of total output} has declined steadily. On a \textit{per capita} basis
Canada's emissions amounted to 23.5 tonnes in 2005, and dropped to 20.3 by
2010. This improvement in efficiency means that Canada's GDP is now \textit{%
less energy intensive}. The quest for increased efficiency is endless, if
economic growth is to continue at rates that will satisfy the world's
citizens and more broadly the impoverished world. The critical challenge is
to produce more output while using not just less energy per unit of output,
but to use less energy in total.

While Canada's energy intensity (GHGs per unit of output) has dropped, our
overall emissions have increased by almost 20\% since 1990. Furthermore,
while developed economies have increased their efficiency, it is the \textit{%
world's} efficiency that is ultimately critical. By outsourcing much of its
manufacturing sector to China, Canada and the West have offloaded some of
their most GHG-intensive activities. But GHGs are a common property resource.

Canada's GHG emissions also have a regional aspect: The \textit{production}
of oil and gas, which has created considerable wealth for all Canadians, is
both energy intensive and concentrated in a limited number of provinces
(Alberta, Saskatchewan and more recently Newfoundland and Labrador).

\newhtmlpage

\subsection*{GHG measurement}

GHG atmospheric concentrations are measured in parts per million (ppm).
Current levels in the atmosphere are at 400 ppm, and continued growth in
concentration will  lead to serious economic and social disruption. In the
immediate pre-industrial revolution era concentrations were in the 250 ppm
range. Hence 500 ppm represents the `doubling' factor that is so frequently
discussed in the media.

GHGs are augmented by the annual additions to the stock already in the
atmosphere, and at the same time they decay---though very slowly.
GHG-reduction strategies that propose an immediate reduction in emissions
are more costly than those aimed at a more gradual reduction. For example, a
slower investment strategy would permit in-place production and
transportation equipment to reach the end of its economic life rather than
be scrapped and replaced `prematurely'. Policies that focus upon longer-term
replacement are therefore less costly in this specific sense.

While not all economists and policy makers agree on the time scale for
attacking the problem, the longer that GHG reduction is postponed, the
greater the efforts will have to be in the long term---because GHGs will
build up more rapidly in the near term.

A critical question in controlling GHG emissions relates to the cost of
their control: How much of annual growth might need to be sacrificed in
order to get emissions onto a sustainable path? Again estimates vary. The
Stern Review (2006) proposed that, with an increase in technological
capabilities, a strategy that focuses on the relative near-term
implementation of GHG reduction measures might cost
``only'' a few percentage points of the value of world
output. If correct, this is a low price to pay for risk avoidance in the
longer term.

Nonetheless, such a reduction will require particular economic policies, and
specific sectors will be impacted more than others.

\newhtmlpage

\subsection*{Economic policies for climate change}

There are three main ways in which polluters can be controlled. One involves
issuing direct controls; the other two involve incentives---in the form of
pollution taxes, or on tradable ``permits'' to pollute.

To see how these different policies operate, consider first Figure~\ref{fig:optimalpollution}.
It is a standard diagram in environmental economics,
and is somewhat similar to our supply and demand curves. On the horizontal
axis is measured the quantity of environmental damage or pollution, and on
the vertical axis its dollar value or cost. The upward-sloping damage curve
represents the cost to society of each additional unit of pollution or gas,
and it is therefore called a \terminology{marginal damage curve}. It is
positively sloped to reflect the reality that, at low levels of emissions,
the damage of one more unit is less than at higher levels. In terms of our
earlier discussion, this means that an increase in GHGs of 10 ppm when
concentrations are at 300 ppm may be less damaging than a corresponding
increase when concentrations are at 500 ppm.

\begin{DefBox}
The \textbf{marginal damage curve} reflects the cost to society of an additional unit of pollution.
\end{DefBox}


% Figure 5.7	fig:optimalpollution
\input{figures/ch5/ch5fig7}

\newhtmlpage

The second curve is the abatement curve. It reflects the cost of reducing
emissions by one unit, and is therefore called a \terminology{marginal
abatement curve}. This curve has a negative slope indicating that, as we
reduce the total quantity of pollution produced, the cost of further unit
reductions rises. This shape corresponds to reality. For example, halving
the emissions of pollutants and gases from automobiles may be achieved by
adding a catalytic converter and reducing the amount of lead in gasoline.
But reducing those emissions all the way to zero requires the development of
major new technologies such as electric cars---an enormously more costly
undertaking.

\begin{DefBox}
The \textbf{marginal abatement curve} reflects the cost to society of reducing the quantity of pollution by one unit.
\end{DefBox}

If producers are unconstrained in the amount of pollution they produce, they
will produce more than what we will show is the optimal amount --
corresponding to $Q^*$. This amount is optimal in the sense that at
levels greater than $Q^*$ the damage exceeds the cost of reducing the
emissions. However, reducing emissions below $Q^*$ would
mean incurring a cost per unit reduction that exceeds the benefit of that
reduction. Another way of illustrating this is to observe that at a level of
pollution above $Q^*$ the cost of reducing it is less than the damage
it inflicts, and therefore a net gain accrues to society as a result of the
reduction. But to reduce pollution below $Q^*$ would involve an
abatement cost greater than the reduction in pollution damage and therefore
no net gain to society. This constitutes a first rule in optimal pollution
policy.

\textit{An optimal quantity of pollution occurs when the marginal cost of
abatement equals the marginal damage.}

A second guiding principle emerges by considering a situation in which some
firms are relatively `clean' and others are `dirty'. More specifically, a
clean firm A may have already invested in new equipment that uses less
energy per unit of output produced, or emits fewer pollutants per unit of
output. In contrast, the dirty firm B uses older dirtier technology. Suppose
furthermore that these two firms form a particular sector of the economy and
that the government sets a limit on total pollution from this sector, and
that this limit is less than what the two firms are currently producing.
What is the least costly method to meet the target?

\newhtmlpage

The intuitive answer to this question goes as follows: In order to reduce
pollution at least cost to the sector, calculate what it would cost each
firm to reduce pollution from its present level. Then implement a system so
that the firm with the least cost of reduction is the first to act. In this
case the `dirty' firm will likely have a lower cost of abatement since it
has not yet upgraded its physical plant. This leads to a second rule in
pollution policy:

\textit{With many polluters, the least cost policy to society requires
producers with the lowest abatement costs to act first.}

This principle implies that policies which impose the same emission limits
on firms may not be the least costly manner of achieving a target level of
pollution. Let us now consider the use of \terminology{tradable permits} and %
\terminology{corrective/carbon taxes} as policy instruments. These are
market-based systems aimed at reducing GHGs.

\begin{DefBox}
\textbf{Tradable permits} and \textbf{corrective/carbon taxes} are market-based systems aimed at reducing GHGs.
\end{DefBox}

\newhtmlpage

\subsection*{Incentive mechanism I: Tradable permits}

A system of tradable permits is frequently called a `cap and trade' system,
because it limits or caps the total permissible emissions, while at the same
time allows a market to develop in permits. For illustrative purposes,
consider the hypothetical two-firm sector we developed above, composed of
firms A and B. Firm A has invested in clean technology, firm B has not. Thus
it is less costly for B to reduce emissions than A if further reductions are
required. Next suppose that each firm is allocated by the government a
specific number of `GHG emission permits'; and that the total of such
permits is less than the amount of emissions at present, and that each firm
is emitting more than its permits allow. How can these firms achieve the
target set for this sector of the economy?

The answer is that they should be able to engage in mutually beneficial
trade: If firm B has a lower cost of reducing emissions than A, then it may
be in A's interest to pay B to reduce B's emissions heavily. Imagine that
each firm is emitting 60 units of GHG, but they have permits to emit only 
50 units each. And furthermore suppose it costs B \$20 to reduce GHGs by
one unit, whereas it costs A \$30 to do this. In this situation A could
pay B \$25 for several permits and this would benefit both firms. B can
reduce GHGs at a cost of \$20 and is being paid \$25 to do this. In turn
A would incur a cost of \$30 per unit to reduce his GHGs but he can buy
permits from B for just \$25 and avoid the \$30 cost. Both firms gain,
and the total cost to the economy is lower than if each firm had to reduce
by the same amount.

The benefit of the cap 'n trade system is that it enables the marketplace to
reduce GHGs at least cost.

The largest system of tradable permits currently operates in the European
Union: The \textit{EU Emissions Trading System}. It covers more than 10,000
large energy-using installations. Trading began in 2005. In North America a
number of Western states and several Canadian provinces are joined, either
as participants or observers, in the \textit{Western Climate Initiative},
which is committed to reduce GHGs by means of tradable emissions permits.
The longer-term goal of these systems is for the government to issue
progressively fewer permits each year, and to include an ever larger share
of GHG-emitting enterprises with the passage of time.

\newhtmlpage

\subsection*{Policy in practice -- international}

In an ideal world, permits would be traded internationally, and such a
system might be of benefit to developing economies: If the cost of reducing
pollution is relatively low in developing economies because they have few
controls in place, then developed economies, for whom the cost of GHG
reduction is high, could induce firms in the developing world to undertake
cost reductions. Such a trade would be mutually beneficial. For example,
imagine in the above example that B is located in the developing world and A
in the developed world. Both would obviously gain from such an arrangement,
and because GHGs are a common property, the source of GHGs from a damage
standpoint is immaterial.

\subsection*{Incentive mechanism II: Taxes}

Corrective taxes are frequently called \textit{Pigovian} taxes, after the
economist Arthur Pigou. He advocated taxing activities that cause negative
externalities. These taxes have been examined above in Section~\ref{sec:ch5sec4}.
Corrective taxes of this type can be implemented as part of
a tax package \textit{reform}. For example, taxpayers are frequently
reluctant to see governments take `yet more' of their money, in the form of
new taxes. Such concerns can be addressed by reducing taxes in other sectors
of the economy, in such a way that the package of tax changes maintains a
`revenue neutral' impact.

\newhtmlpage

\subsection*{Revenues from taxes and permits}

Taxes and tradable permits differ in that taxes generate revenue for the
government from polluting producers, whereas permits may not generate
revenue, or may generate less revenue. If the government simply \textit{%
allocates} permits initially to all polluters, free of charge, and allows a
market to develop, such a process generates no revenue to the government.
While economists may advocate an \textit{auction} of permits in the start-up
phase of a tradable permits market, such a mechanism may run into political
objections.

Setting taxes at the appropriate level requires knowledge of the cost and
damage functions associated with GHGs. At the present time, economists
and environmental scientists think that an appropriate price or tax on one
tonne of GHG is in the $\$20-\$40$ range. Such a tax would reduce emissions
to a point where the longer-term impact of GHGs would not be so severe as
otherwise.

British Columbia introduced a carbon tax of \$10 per tonne of GHG on fuels
in 2008, and has increased that price regularly. It stood at \$30 per tonne
in 2012. This tax was designed to be revenue neutral in order to make it
more acceptable. This means that British Columbia reduced its income tax
rates by an amount such that income tax payments would fall by an amount
equal to the revenue captured by the carbon tax. A recent study in British 
Columbia recommended that the current \$30-per-tonne carbon tax be increased 
to a multiple of its present level over the coming decades.
\footnote{Climate Leadership Team, Recommendations to Government: \url{http://goo.gl/6gjvSc}}

\subsection*{Policy in practice -- domestic large final emitters}

Governments frequently focus upon quantities emitted by individual firms,
perhaps because governments are reluctant to introduce carbon taxes or a
system of tradable permits. Specifically the focus is upon firms called 
\textit{large final emitters} (LFEs). Frequently, a relatively small number
of producers are responsible for a disproportionate amount of an economy's
total pollution, and limits are placed on those firms in the belief that
significant economy-wide reductions can be achieved in this manner. A
further reason for concentrating on these LFEs is that the monitoring costs
are relatively small compared to the costs associated with monitoring 
\textit{all} firms in the economy. It must be kept in mind that pollution
permits may be a legal requirement in some jurisdictions, but monitoring is
still required, because firms could choose to risk polluting without owning
a permit.

\section*{Conclusion}

Welfare economics lies at the heart of public policy. Demand and supply
curves can be interpreted as value curves and cost curves when there are no
externalities involved. This is what enables us to define an efficient
output of a product, and consequently an efficient use of the economy's
resources. While efficiency is a central concept in economics, we must keep
in mind that when the economic environment changes so too will the efficient
use of resources, as we illustrated in Section~\ref{sec:ch5sec7}.

In this chapter we have focused on equity issues through the lens of GHG
emissions. The build-up of GHGs in our atmosphere invokes the concept of 
\textit{intergenerational equity}: The current generation is damaging the
environment and the costs of that damage will be borne by subsequent
generations. Hence it is inequitable in the intergenerational sense for us
to leave a negative legacy to succeeding generations. Equity arises within
generations also. For example, how much more in taxes should the rich pay
relative to the non-rich? We will explore this type of equity in our chapter
on government.
